# Two-stage cluster random sampling {#Twostage}

As opposed to cluster random sampling in which all population units of cluster are observed (Chapter \@ref(Cl)), in two-stage cluster random sampling\index{Two-stage cluster random sampling} not all sampling units of the selected clusters are observed, but only some of them. In two-stage cluster random sampling the clusters will generally be contiguous groups of units, for instance all points in a map polygon (the polygons on the map are the clusters), whereas in single-stage cluster random sampling the clusters generally are non-contiguous. The sampling units to be observed are selected by random subsampling of the randomly selected clusters. In two-stage cluster sampling the clusters are commonly referred to as primary sampling units\index{Primary sampling units} (psu's) or shortly primary units (pu's), and the units selected in the second stage as the secondary sampling units\index{Secondary sampling units} (ssu's) or secondary units (su's).

As with cluster random sampling, two-stage cluster random sampling may lead to a strong spatial clustering of the selected sampling units (ssu's) in the study area. This may save considerable time for fieldwork, and more population units can be observed for the same budget. However, due to the spatial clustering the estimates will generally be less precise compared to samples of the same size selected by a design that leads to a much better spreading of the sampling units throughout the study area, such as systematic random sampling.

In two-stage cluster random sampling in principle any type of sampling design can be used at the two stages, leading to numerous combinations. An example is (SI,SI), in which both psu's and ssu's are selected by simple random sampling.

Commonly the psu's have unequal size, i.e. the number of ssu's (finite population) or the area (infinite population) are not equal for all psu's. Think for instance of the agricultural fields, forest stands, lakes, river sections etc., in an area. If the psu's are of unequal size, then psu's can best be selected with probabilities proportional to their size (pps). Recall that in (one-stage) cluster random sampling I also recommended to select the clusters with probabilities proportional to their size, see Chapter \@ref(Cl). If the total of the study variable of a psu is proportional  to its size, then pps sampling leads to more precise estimates compared to simple random sampling of psu's. Also, with pps sampling of psu's the estimation of means or totals and of their sampling variances is much simpler compared to selection with equal probabilities. Implementation of selection with probabilities proportional to size is easiest when units are replaced (pps with replacement, ppswr)\index{pps sampling!with replacement (ppswr)}. This implies that a psu might be selected more than once, especially if the total number of psu's in the population is small compared to the number of psu draws (large sampling fraction in first stage).
 
Using a list as a sampling frame, the following algorithm can be used to select $n$ times a psu by ppswr from a total of $N$ psu's in the population:

1. Select randomly one ssu from the list with $M=\sum_{j=1}^N M_j$ ssu's ($M_j$ is number of ssu's of psu $j$), and determine the psu of the selected ssu.  
2. Repeat step 1 until $n$ selections have been made.  

In the first stage a ssu is selected in order to select a psu. This may seem unnecessarily complicated. The reason for this is that this procedure automatically adjusts for the size of the psu's (number of ssu's within a psu), i.e. a psu is selected with probability proportional to its size. In the second stage, a pre-determined number of secondary sampling units, $m_{j}$, is selected every time psu $j$ is selected.

Note that the ssu selected in the first step of the two algorithms primarily serve to identify the psu, but these ssu's can also be used as selected ssu's.   

The selection of a two-stage cluster random sample is illustrated again with Voorst. Twenty-four blocks of 0.5 km $\times$ 0.5 km are constructed that serve as psu's.  

```{block2, type='rmdnote'}
Due to built-up areas, roads etc., the psu's in Voorst have unequal size, i.e. the number of secondary units (pixels) within the psu's varies among the psu's.
```


```{r}
load("data/Voorst.RData")
w <- 500 #width of psu's
s1bnd <- seq(from=min(grdVoorst$s1)+w, to=min(grdVoorst$s1)+(11*w), by=w)+12.5
s1f <- findInterval(grdVoorst$s1,s1bnd)
s2bnd <- min(grdVoorst$s2)+w+12.5
s2f <- findInterval(grdVoorst$s2, s2bnd)
grdVoorst$psu <- as.character(interaction(s1f, s2f))
```

To select a two-stage cluster random sample a function is defined.

```{r}
twostage <- function(sframe, psu, n, m) {
  units <- sample.int(nrow(sframe), size=n, replace=TRUE)
  mypsusample <- sframe[units,psu]
  ssunits <- NULL
  for (psunit in mypsusample) {
    ssunit <- sample(x = which(sframe[,psu]==psunit), size=m, replace=TRUE)
    ssunits <- c(ssunits, ssunit)
  }
  psudraw <- rep(c(1:n), each=m)
  mysample <- data.frame(ssunits,sframe[ssunits,], psudraw)
  mysample
}
```

Note that both the primary and secondary sampling units are selected with replacement. The secondary units are selected by simple random sampling with replacement because the actual population is infinite. The infinite population is discretised by a finite number of grid nodes (centers of grid cells). If a grid node is selected, one point is selected fully randomly from the associated grid cell. This is done with function `jitter`. In every grid cell there is an infinite number of points, so we must select the grid cells with replacement. If a grid node is selected more than once, more than one point is selected from that grid cell. The column `psudraw` in the output data frame of function `twostage` is needed in estimation because psu's are selected with replacement. In case a psu is selected more than once, multiple estimates of the mean of that psu are used in estimation, see next section.

The function `twostage` is used to select four times a psu ($n=4$), with probabilities proportional to size and with replacement (ppswr). The second stage sample size $m_i$ equals 10 for all psu's. These secondary sampling units (ssu's) are selected by simple random sampling. 

```{r}
n <- 4
m <- 10
set.seed(314)
mysample <- twostage(sframe=grdVoorst, psu="psu", n=n, m=m)
cellsize <- 25
mysample$s1 <- jitter(mysample$s1, amount=cellsize/2)
mysample$s2 <- jitter(mysample$s2, amount=cellsize/2)
```

Figure \@ref(fig:TwostageVoorst) shows the selected sample.

```{r TwostageVoorst, echo=FALSE, out.width='100%', fig.cap="Two-stage cluster random sample from Voorst. Four times a psu is selected with probabilities proportional to their size. Each time a psu is selected, ten points are selected from that psu by simple random sampling."}
ggplot(data=grdVoorst) +        
  geom_raster(mapping = aes(x = s1/1000, y = s2/1000, fill=as.factor(psu))) +
  scale_fill_viridis_d(alpha=0.5) +
  geom_point(data=mysample, mapping = aes(x=s1/1000, y = s2/1000), size = 2) +
  scale_x_continuous(name = "Easting (km)") +
  scale_y_continuous(name = "Northing (km)") +
  coord_fixed()+
  theme(legend.position = "none")
```

## Estimation of population parameters {#twostagesamplingestimators}

The population total can be estimated by substituting the estimated cluster (primary unit) totals in Equation \@ref(eq:EstTotalCl). This yields the following estimator for the population total:

\begin{equation}
\hat{t}(z) = \frac{M}{n} \sum_{j \in \mathcal{S}} \frac{\hat{t}_{j}(z)}{M_{j}} = \frac{M}{n} \sum_{j \in \mathcal{S}} \hat{\bar{z}}_{j} \;,
(\#eq:EstTotalTwostage)
\end{equation}

where $n$ is the number of psu selections and $M_{j}$ is total number of ssu's in psu $j$. This shows that the mean of cluster $j$, $\bar{z}_j$, is replaced by the estimated mean of psu $j$, $\hat{\bar{z}}_j$. Dividing this estimator by the total number of population units $M$ gives the pwr  estimator of the population mean:

\begin{equation}
\hat{\bar{\bar{z}}}=
\frac{1}{n}\sum\limits_{j \in \mathcal{S}}\hat{\bar{z}}_{j} \;,
(\#eq:EstMeanTwostage)
\end{equation}

with $\hat{\bar{z}}_{j}$ the estimated mean of the psu $j$. With simple random sampling of ssu's this mean can be estimated by the sample mean of this psu. Note the two bars $\hat{\bar{\bar{z}}}$, indicating that the population mean is estimated as the mean of estimated primary unit means. When $m_i$ is equal for all psu's the sampling design is self-weighting\index{Self-weighting sampling design}, i.e. the average of $z$ over all selected secondary units is an unbiased estimator of the population mean.

The sampling variance of the estimator of the mean with two-stage cluster random sampling (primary units selected with probabilities proportional to size with replacement, secondary units by simple random sampling (with replacement in case of finite populations) and $m_j = m, j = 1, \dots, N$) is equal to (@coc77, Equation (11.33))

```{block2, type='rmdnote'}
The equation in Cochran (1977) is the variance estimator for the population total. In Exercise 5 you are asked to derive the estimator of the variance of the estimator of the population mean from the estimator of the variance of the estimator of the population total.
```

\begin{equation}
V(\hat{\bar{\bar{z}}}) = \frac{S^2_{\mathrm{b}}}{n} + \frac{S^2_{\mathrm{w}}}{n\;m} \;,
(\#eq:TrueVarEstMeanTwostage)
\end{equation}

with

\begin{equation}
S^2_{\mathrm{b}}=\sum_{j=1}^N p_j\left(\bar{z}_j-\bar{z}\right)^2\;,
(\#eq:PooledBetweenClusterVariance)
\end{equation}

and

\begin{equation}
S^2_{\mathrm{w}}=\sum_{j=1}^N p_j S^2_j
(\#eq:PooledWithinClusterVariance)
\end{equation}

with $N$ the total number of psu's in the population, $p_j=M_j/M$ the draw-by-draw selection probability of psu $j$\index{Draw-by-draw selection probability!of a primary sampling unit}, $\bar{z}_j$ the mean of psu $j$, $\bar{z}$ the population mean of $z$, and $S^2_j$ the variance of $z$ within primary unit $j$:

\begin{equation}
S^2_j = \frac{1}{M_j} \sum_{k=1}^{M_j} (z_{kj}-\bar{z}_j)^2 \;.
(\#eq:WithinClusterVariance)
\end{equation}

```{block2, type='rmdnote'}
The first term of Equation \@ref(eq:TrueVarEstMeanTwostage) is equal to the variance of Equation \@ref(eq:TrueVarEstMeanCl). This variance component accounts for the variance of the true primary unit means within the population. The second variance component quantifies our additional uncertainty about the population mean, as we do not observe all secondary units of the selected primary units, but only a subset (sample) of these units.
```

The sampling variance of the estimator of the population mean can simply be estimated by

\begin{equation}
\widehat{V}\!\left(\hat{\bar{\bar{z}}}\right)=\frac{\widehat{S^2}(\hat{\bar{z}})}{n}  \;,
(\#eq:VarEstMeanTwostage)
\end{equation}

with $\widehat{S^2}(\hat{\bar{z}})$ the estimated variance of the *estimated* primary unit means:

\begin{equation}
\widehat{S^2}(\hat{\bar{z}}) = \frac{1}{n-1}\sum_{j \in \mathcal{S}}(\hat{\bar{z}}_{j}-\hat{\bar{\bar{z}}})^2 \;,
(\#eq:S2psuMeans)
\end{equation}

with $\hat{\bar{z}}_{j}$ the estimated mean of psu $j$, and $\hat{\bar{\bar{z}}}$ the estimated population mean (Equation \@ref(eq:EstMeanTwostage)). 

```{block2, type='rmdnote'}
Neither the sizes of the psu's, $M_j$, nor the secondary sample sizes $m_{j}$ occur in Equations \@ref(eq:VarEstMeanTwostage) and \@ref(eq:S2psuMeans). This simplicity is due to the fact that the psu's are selected with replacement and with probabilities proportional to size. The effect of the secondary sample sizes on the variance is implicitly accounted for. To understand this, note that the larger $m_{j}$, the less variable $\hat{\bar{z}}_{j}$, and the smaller its contribution to the variance.
```

Let us assume a linear model for the total costs: $C = c_0 + c_1n + c_2n\;m$, with $c_0$ the fixed costs, $c_1$ the cost per primary unit and $c_2$ the cost per secondary unit. We want to minimise the total costs, under the constraint that the variance of the estimator of the population mean may not exceed $V_{\mathrm{max}}$. The total costs can then be minimised by selecting [@gru06]

\begin{equation}
n=\frac{1}{V_{\mathrm{max}}}\left(S_{\mathrm{w}}S_{\mathrm{b}}\sqrt{\frac{c_2}{c_1}}+S^2_{\mathrm{b}}\right)
(\#eq:nopt)
\end{equation}

primary units, and 

\begin{equation}
m=\frac{S_{\mathrm{w}}}{S_{\mathrm{b}}}\sqrt{\frac{c_1}{c_2}}
(\#eq:mopt)
\end{equation}

secondary units per primary unit.

Conversely, given a budget $C_{\mathrm{max}}$, the optimal number of primary unit selections\index{Optimal sample size in two-stage cluster random sampling} can be computed with [@gru06]

\begin{equation}
n=\frac{C_{\mathrm{max}}S_{\mathrm{b}}}{S_{\mathrm{w}}\sqrt{c_1c_2}+S_{\mathrm{b}}c_1}\;,
(\#eq:nopt2)
\end{equation}

and $m$ as above.

In **R** the population mean and the sampling variance of the estimator of the mean can be estimated as follows.

```{r}
mz_psu <- tapply(mysample$z, INDEX=mysample$psudraw, FUN=mean)
mz <- mean(mz_psu)
se_mz <- sqrt(var(mz_psu)/n)
```

The estimated mean equals `r round(mz,3)` and the estimated standard error equals `r round(se_mz,3)`. The sampling design is self-weighting, and so the estimated mean is equal to the sample mean.

```{r}
print(mean(mysample$z))
```

The same estimate is obtained with functions `svydesign` and `svymean` of package **survey** [@Lumley2020]. The estimator of the population total can be written as a weighted sum of the observations with all weights equal to $M/(n\;m$. These weights are assigned to argument `weight`. 

```{r}
library(survey)
M <- nrow(grdVoorst)
mysample$weights <- M/(n*m)
design_2stage <- svydesign(id=~psudraw+ssunits, weight=~weights, data=mysample)
svymean(~z, design_2stage, deff="replace")
```

Similar to (one-stage) cluster random sampling the estimated design effect\index{Design effect} is much larger than 1.

A confidence interval estimate of the population mean can be computed with method `confint`. The number of degrees of freedom equals the number of psu draws minus one.

```{r}
confint(svymean(~z, design_2stage, df=degf(design_2stage), level=0.95))
```

Figure \@ref(fig:SamplingDistributionTwostage) shows the sampling distributions of the pwr estimator of the population mean with two-stage cluster random sampling and of the $\pi$ estimator with simple random sampling from Voorst, obtained by repeating the random sampling with each design and estimation 10,000 times. For simple random sampling the sample size is equal to $n \times m$. 

```{r, echo = FALSE, eval=FALSE}
number_of_samples <- 10000
mz <- v_mz <- mz_SI <- numeric(length=number_of_samples)

set.seed(31415)
SI <- function(sframe, n) {
    units <- sample.int(nrow(sframe), size=n, replace=FALSE)
    mysample <- sframe[units,]
    mysample
}

for (i in 1:number_of_samples) {
  mysample <- twostage(grdVoorst, "psu", n, m)
  psuMeans <- tapply(mysample$z, INDEX=mysample$psudraw, FUN=mean)
  mz[i] <- mean(psuMeans, na.rm=TRUE)
  v_mz[i] <- var(psuMeans, na.rm=TRUE)/n
  mySIsample <- SI(grdVoorst, n=n*m)
  mz_SI[i] <- mean(mySIsample$z)
}

save(mz, mz_SI, v_mz, file="results/Twostage_Voorst.RData")
```

```{r SamplingDistributionTwostage, echo=FALSE, fig.asp=.8, fig.cap="Sampling distribution of pwr estimator of mean of SOM (g/kg) in Voorst with two-stage random sampling, and of the $\\pi$ estimator with simple random sampling, bot design with a sample size of 40 units."}
load(file="results/Twostage_Voorst.RData")
estimates <- data.frame(mz, mz_SI)
names(estimates) <- c("Twostage","SI")
df <- estimates %>% pivot_longer(.,cols=c("Twostage","SI"))
df$name <- factor(df$name, levels=c("Twostage","SI"), ordered=TRUE )
ggplot(data = df) +
    geom_boxplot(aes(y=value, x=name)) +
    geom_hline(yintercept=mean(grdVoorst$z), colour="red")+
    scale_x_discrete(name = "Sampling design") +
    scale_y_continuous(name = "Estimated mean SOM")
v_mz_sim <- var(mz)
m_vmz_sim <- mean(v_mz)
v_mz_SI_sim <- var(mz_SI)
```

The variance of the 10,000 means with two-stage cluster random sampling equals `r as.character(round(var(mz),3))`. This is considerably larger than with simple random sampling: `r as.character(round(var(mz_SI),3))`. The average of the estimated variances equals `r as.character(round(mean(v_mz),3))`. 

Optimal sample sizes for two-stage cluster random sampling (ppswr in first stage, simple random sampling without replacement in second stage) can be computed with function `clusOpt2` of **R** package **PracTools** (@PracTools, @Vaillant2018). This function requires as input various variance measures, which can be computed with function `BW2stagePPS` in case the study variable is known for the whole population, or estimated from a sample with function `BW2stagePPSe`. This is left as an exercise (Exercise 5).

#### Exercises {-}

1. Write an **R** script to compute the true sampling variance of the estimator of the population mean for $n=4$ and $m=10$, see Equation \@ref(eq:TrueVarEstMeanTwostage).   
2. Do you expect that the standard error of the estimated population mean with ten psu draws ($n=10$) and four ssu's per psu draw ($m=4$) are larger or smaller than with four psu draws ($n=4$) and ten ssu's per psu draw ($m=10$)?  
3. Compute the optimal sample sizes $n$ and $m$ for a maximum variance of the estimator of the population mean of 1, $c_1=2$ and $c_2=1$ monetary unit, see Equations \@ref(eq:nopt) and \@ref(eq:mopt).  
4. Compute the optimal sample sizes $n$ and $m$ for a budget of 100 monetary units,  $c_1=2$ and $c_2=1$ monetary units, see Equations \@ref(eq:nopt2) and \@ref(eq:mopt).  
5. Use function `clusOpt2` of **R** package **PracTools** to compute optimal sample sizes given the precision requirement for the estimated population mean of Exercise 3 and given the budget of Exercise 4. First use function `BW2stagePPS` to compute the variance measures needed as input for function ` optClus2`. Note that the precision requirement of function `clusOpt2` is the coefficient of variation of the estimated population total, i.e. the standard deviation of the estimated population total divided by the population total. Compute this coefficient of variation from the maximum variance of the estimator of the population mean used in Exercise 3.  
6. Derive the variance estimator for the estimated population mean, Equation \@ref(eq:TrueVarEstMeanTwostage), from the variance estimator of the estimated population total [@coc77]:
\begin{equation}
V(\hat{t}(z)) = \frac{1}{n} \sum_{j=1}^N p_j\left(\frac{t_j(z)}{p_j}-t(z)\right)^2 + \frac{1}{n} \sum_{j=1}^N \frac{M_j^2 (1-f_{2j})S^2_j}{m_j p_j}  \;,
(\#eq:EstVarTotalTwostageCochran)
\end{equation}
with $\hat{t}(z)$ and $t(z)$ the estimated and true population total of $z$, respectively, $t_j(z)$ the total of psu $j$, and $p_j = M_j/M$. Use $m_j = m, j = 1, \dots, N$, and $f_{2j}=0$, i.e. sampling from infinite population, or sampling of ssu's within psu's by simple random sampling *with* replacement from finite population.

## Primary sampling units selected with probabilities proportional to size, without replacement

Similar to cluster random sampling, we may prefer to select the primary sampling units without replacement\index{pps sampling!without replacement (ppswor)}. This leads to less strong spatial clustering of the sampling points, especially with large sampling fractions of primary sampling units. Sampling without replacement of psu's can be done with function `UPpivotal` of package **sampling** [@Tille2016], see Section \@ref(pivotalmethod). The second stage sample of secondary sampling units is selected with function `strata` of the same package, using the psu's as strata.

```{r}
library(sampling)
M_psu <- tapply(grdVoorst$z, INDEX=grdVoorst$psu, FUN=length)
n <- 6
pi <- n*M_psu/M
set.seed(314)
sampleind <- UPpivotal(pik=pi, eps=1e-6)
psus <- sort(unique(grdVoorst$psu))
sampledpsus <- psus[sampleind==1]
mysample_stage1 <- grdVoorst[grdVoorst$psu %in% sampledpsus,]
units <- sampling::strata(mysample_stage1, stratanames="psu",
  size=rep(m,n), method="srswor")
mysample <- getdata(mysample_stage1, units)
mysample$ssunits <- units$ID_unit
mysample$pi <- n*m/M
print(mean_HT <- sum(mysample$z/mysample$pi)/M)
```

The population mean can be estimated with function `svymean` of package **survey**. To estimate the variance, a simple solution is to treat the two-stage cluster random sample as a pps sample *with replacement*, so that variance can be estimated with Equation \@ref(eq:VarEstMeanTwostage). With small sampling fractions of psu's the overestimation of the variance is negligible. With larger sampling fractions Brewer's method is recommended,  see @Berger2004 (option 2)\index{Brewer's variance estimator}.

```{r}
mysample$fpc1 <- n*M_psu[mysample$psu]/M
mysample$fpc2 <- m/M_psu[mysample$psu]
design_2stageppswor <- svydesign(id=~psu+ssunits, data=mysample,
  pps="brewer", fpc=~fpc1+fpc2)
svymean(~z,design_2stageppswor)
```


## Simple random sampling of primary sampling units

Suppose the primary sampling units are for some reason not selected  with probabilities proportional to their size, but by simple random sampling without replacement. The inclusion probability of the psu's then equal $\pi_j=n/N, j = 1, \dots, N$, and the population total can be estimated by (compare with Equation \@ref(eq:EstTotalClEqual))

\begin{equation}
\hat{t}(z) =  \sum_{j=1}^n \frac{\hat{t}_j(z)}{\pi_j} = \frac{N}{n} \sum_{j=1}^n \hat{t}_j(z)\;,
(\#eq:EstTotalTwostageEqual)
\end{equation}

with $\hat{t}_j(z)$ an estimator of the total of psu $j$. The population mean can be estimated by dividing this estimator by the population size $M$. 

Alternatively, we may estimate the population mean by dividing the estimate of the population total by the *estimated* population size. The population size can be estimated by the $\pi$ estimator, see Equation \@ref(eq:EstPopulatonSizeClEqual). The $\pi$ estimator and ratio estimator are equal when the psu's are selected by ppswr, but not so when the psu's of different size are selected with equal probabilities. This is shown below. First a sample is selected by selecting both psu's and ssu's by simple random sampling without replacement. 

```{r}
library(sampling)
set.seed(314)
psus <- sort(unique(grdVoorst$psu))
ids_psu <- sample.int(length(psus), size=n, replace=FALSE) 
sampledpsus <- psus[ids_psu]
mysample_stage1 <- grdVoorst[grdVoorst$psu %in% sampledpsus,]
units <- sampling::strata(mysample_stage1, stratanames="psu",
  size=rep(m,n), method="srswor")
mysample <- getdata(mysample_stage1, units)
mysample$ssunits <- units$ID_unit
```

The population mean is estimated by the $\pi$ estimator and the ratio estimator.

```{r}
N <- length(unique(grdVoorst$psu))
M_psu <- tapply(grdVoorst$z, INDEX=grdVoorst$psu, FUN=length)
pi_psu <- n/N
pi_ssu <- m/M_psu[mysample$psu]
mysample$pi <- pi_psu*pi_ssu
z_piexpanded <- with(mysample,z/pi)
tz_HT <- sum(z_piexpanded)
mz_HT <- tz_HT/M
M_HT <- sum(1/mysample$pi)
mz_ratio <- tz_HT/M_HT
```

The $\pi$ estimate equals `r round(mz_HT,3)` and the ratio estimate equals `r round(mz_ratio,3)`. The $\pi$ estimate of the population mean can also be computed by first estimating totals of psu's, see Equation \@ref(eq:EstTotalTwostageEqual). 

```{r}
tz_psu <- tapply(mysample$z/pi_ssu, INDEX=mysample$psu, FUN=sum)
tz_HT <- sum(tz_psu/pi_psu)
(mz_HT <- tz_HT/M)
```

The variance of the $\pi$ estimator of the population mean can be estimated by first estimating the variance of the estimator of the psu totals:

\begin{equation}
\widehat{V}(\hat{t}(z)) = N^2\left(1-\frac{n}{N}\right)\frac{\widehat{S^2}(\hat{t}_i(z))}{n}
\;,
(\#eq:EstVarTotalHTTwostageSI)
\end{equation}

and dividing this variance by the squared number of population units:

\begin{equation}
\widehat{V}(\hat{\bar{\bar{z}}}) = \frac{1}{M^2} \widehat{V}(\hat{t}(z))
\;.
(\#eq:EstVarMeanHTTwostageSI)
\end{equation}


```{r}
fpc <- 1-n/N
v_tz <- N^2*fpc*var(tz_psu)/n
(se_mz_HT <- sqrt(v_tz/M^2))
```

The ratio estimator of the population mean and its standard error can be computed with function `svymean` of package **survey**.

```{r}
mysample$fpc1 <- N
mysample$fpc2 <- M_psu[mysample$psu]
design_2stage <- svydesign(id=~psu+ssunits, fpc=~fpc1+fpc2, data=mysample)
svymean(~z, design_2stage)
```
The estimated standard error of the ratio estimator is slightly smaller than the standard error of the $\pi$ estimator.


## Stratified two-stage cluster random sampling {#StratifiedTwostage}

The basic sampling designs stratified random sampling (Chapter \@ref(STSI)) and two-stage cluster random sampling can be combined into stratified two-stage cluster random sampling\index{Stratified random sampling!stratified two-stage cluster random sampling}. Figure \@ref(fig:STtwostage) shows a stratified two-stage cluster random sample from Voorst. The strata are 2 km $\times$ 1 km blocks, as before in stratified cluster random sampling (Figure \@ref(fig:ClVoorst)). The primary sampling units are 0.5 km $\times$ 0.5 km blocks, as before in (unstratified) two-stage cluster random sampling (Figure \@ref(fig:TwostageVoorst)).  Within each stratum two times a psu is selected by ppswr, and every time a psu is selected, six ssu's (points) are selected by simple random sampling. The stratification avoids the clustering of the selected psu's in one part of the study area. Compared to (unstratified) two-stage cluster random sampling, the geographical spreading of the psu's is somewhat improved, which may lead to an increase of the precision of the estimated population mean.

```{r, echo=FALSE}
w <- 1000 #width of strata
s1bnd <- seq(min(grdVoorst$s1)+w, min(grdVoorst$s1)+5*w, w)+12.5
grdVoorst$zonestratum <- as.factor(findInterval(grdVoorst$s1, s1bnd))
levels(grdVoorst$zonestratum) <- rep(c("a","b","c"), each=2)
```


```{r}
n_h <- c(2,3,4)
m <- 6
set.seed(314)
stratumlabels <- unique(grdVoorst$zonestratum)
mysample <- NULL
for (i in 1:3) {
  grd_h <- grdVoorst[grdVoorst$zonestratum==stratumlabels[i],]
  mysample_h <- twostage(sframe=grd_h, psu="psu", n=n_h[i], m=m)
  mysample <- rbind(mysample, mysample_h)
}
mysample$s1 <- jitter(mysample$s1, amount=cellsize/2)
mysample$s2 <- jitter(mysample$s2, amount=cellsize/2)
```

```{r STtwostage, echo=FALSE, out.width='100%', fig.cap="Stratified two-stage random sample from Voorst. Strata are 2 km by 1 km blocks."}
ggplot(grdVoorst) +        
  geom_raster(mapping = aes(x = s1/1000, y = s2/1000, fill=as.factor(psu))) +
  scale_fill_viridis_d(alpha=0.5) +
  geom_point(data=mysample, mapping = aes(x = s1/1000, y = s2/1000), size = 1.5) +
  geom_vline(xintercept=s1bnd[c(2,4)]/1000)+
  scale_x_continuous(name = "Easting (km)") +
  scale_y_continuous(name = "Northing (km)") +
  coord_fixed()+
  theme(legend.position = "none")
```

The population mean can be estimated in much in the same way as with stratified cluster random sampling. With function `svymean` this is an easy task.  

```{r}
N_h <- tapply(grdVoorst$psu, INDEX=grdVoorst$zonestratum,
  FUN=function(x) {length(unique(x))})
M_h <- tapply(grdVoorst$z, INDEX=grdVoorst$zonestratum, FUN=length)
mysample$w1 <- N_h[mysample$zonestratum]
mysample$w2 <- M_h[mysample$zonestratum]
design_str2stage <- svydesign(id=~psudraw+ssunits, strata=~zonestratum,
  weights=~w1+w2, data=mysample, nest=TRUE)
svymean(~z, design_str2stage)
```

```{r, echo=FALSE}
rm(list=ls())
```

