# Regular grid and spatial coverage sampling {#RegularGridSpatialCoverage}

This chapter describes and illustrates two sampling designs by which the sampling locations are evenly spread throughout the study area: regular grid sampling and spatial coverage sampling. In a final section the spatial coverage sampling design is used to fill in the empty spaces of an existing sample.

## Regular grid sampling {#Regulargrid}

Sampling on a regular grid\index{Regular grid} is an attractive option for mapping because of its simplicity. The data collected on the grid nodes are not used for design-based estimation of the population mean or total. For this reason the grid need not be placed randomly on the study area as in systematic random sampling (Chapter \@ref(SY)). The grid can be located such that the grid nodes optimally cover the study area in the sense of the average distance of the nodes of a fine discretisation grid to the nearest node of the sampling grid. Commonly used grid configurations are square and triangular. If the grid data are used in kriging (Chapter \@ref(Introkriging)), the optimal configuration depends, among others, on the semivariogram model. If the study variable shows moderate to strong spatial autocorrelation (see Section \@ref(OrdinaryKriging)), triangular grids outperform square grids.

Besides the shape of the sampling grid cells, we must decide on the grid spacing\index{Grid spacing}. The grid spacing determines the number of sampling units in the study area, i.e. the sample size. There are two options to decide on this spacing, either starting from the available budget or from a requirement on the quality of the map. The latter is explained in Chapter \@ref(MBgridspacing), as this requires a model of the spatial variation, and as a consequence this is an example of model-based sampling. Starting from the available budget and an estimate of the costs per sampling unit, we first compute the affordable sample size. Then we may derive from this number the grid spacing. For square grids, the grid spacing in meters is calculated as $\sqrt{A/n}$, where $A$ is the area in m^2^ and $n$ is the number of sampling units (sample size).

Grids can be selected with function `spsample` of package **sp** [@Pebesma2005]. Argument `offset` is used to select a grid non-randomly. Either a sample size can be specified, using argument `n`, or a grid spacing, using argument `cellsize`. In the next code chunk a square grid is selected with a spacing of 200 m.

```{r grid}
library(sp)
gridded(grdVoorst) <- ~ s1 + s2
mysample <- spsample(
  x = grdVoorst, type = "regular", cellsize = c(200, 200),
  offset = c(0.5, 0.5)) %>% as(., "data.frame")
```

Figure \@ref(fig:gridVoorst) shows the selected square grid.

```{r gridVoorst, echo = FALSE, out.width = '100%', fig.cap = "Non-random square grid sample with a grid spacing of 200 m from Voorst."}
grdVoorst <- as(grdVoorst, "data.frame")
ggplot() +
  geom_raster(data = grdVoorst, mapping = aes(x = s1 / 1000, y = s2 / 1000), fill = "grey") +
  geom_point(data = mysample, mapping = aes(x = x1 / 1000, y = x2 / 1000), size = 1.5) +
  scale_x_continuous(name = "Easting (km)") +
  scale_y_continuous(name = "Northing (km)") +
  coord_fixed() +
  theme(panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_blank())

n <- nrow(mysample)
```

The number of grid points in this example equals `r n`. Nodes of the square grid in parts of the area not belonging to the population of interest, such as built-up areas and roads, are discarded by `spsample` (these nodes are not included in the sampling frame file `grdVoorst`). As a consequence, there are some undersampled areas\index{Undersampled area}, for instance in the middle of the study area where two roads cross. If we use the square grid in spatial interpolation, e.g. by ordinary kriging, we are more uncertain about the predictions in these undersampled areas than in areas where the grid is complete. In the next section I will show how this local undersampling can be avoided.

#### Exercises {-}

1. Write an **R** script to select a square grid of size 100 from West-Amhara in Ethiopia. Use `data.frame` `grdAmhara` of package **sswr** as a sampling frame. Use a fixed starting point of the grid, i.e. do not select the grid randomly.   
    +  Compute the number of selected grid points. How comes that it is not exactly equal to 100?
    +  Select a square grid with a spacing of 10.2 km, and compute the sample size.
    +  Write a for-loop to select 200 times a square grid of, on average, 100 points with random starting point. Set a seed so that the result can be reproduced. Determine for each randomly selected grid the number of selected grid points, and save this in a numeric. Compute summary statistics of the sample size, and plot a histogram.
    +  Select a square grid of exactly 100 points.  
  
## Spatial coverage sampling {#SpatialCoverage}

Local undersampling with regular grids can be avoided by relaxing the constraint that the sampling units are restricted to the nodes of a regular grid. This is what is done in *spatial coverage sampling*\index{Spatial coverage sampling} or, in case of a sample that is added to an existing sample, in *spatial infill sampling*\index{Spatial infill sampling}. Spatial coverage and infill samples cover the area or fill in the empty space as uniformly as possible. The sampling units are obtained by minimising a criterion that is defined in terms of the geographic distances between the nodes of a fine discretisation grid and the sampling units. @bru07c proposed to minimise the mean of the squared distances of the grid nodes to their nearest sampling unit (mean squared shortest distance, MSSD\index{Mean squared shortest distance}):

\begin{equation}
MSSD=\frac{1}{N}\sum_{k=1}^{N}\min_{j}\left(D_{kj}^{2}\right) \;,
(\#eq:MSSD)
\end{equation}

where $N$ is the total number of nodes of the discretisation grid and $D_{kj}$ is the distance between the $k$th grid node and the $j$th sampling point. This distance measure can be minimised by the k-means algorithm, which is a numerical, iterative procedure. Figure \@ref(fig:spatialcoveragesamplefromsquare) illustrates the selection of a spatial coverage sample of four points from a square. In this simple example the optimal spatial coverage sample is known, being the centres of the four subsquares of equal size. A simple random sample of four points serves as the initial solution. Each raster cell is then assigned to the closest sampling point. This is the initial clustering. In the next iteration the centres of the initial clusters are computed. Next, the raster cells are re-assigned to the closest new centres. This continues until there is no change anymore. In this case only nine iterations are needed, where an iteration consists of computing the clusters by assigning the raster cells to the nearest centre (sampling unit), followed by computing the centres of these clusters. Figure \@ref(fig:spatialcoveragesamplefromsquare) shows the first, second, and ninth iteration. 

```{r spatialcoveragesamplefromsquare, echo = FALSE, fig.show = 'hold', out.width = '47%', fig.cap = "First, second, and ninth iteration of the k-means algorithm to select a spatial coverage sample of four points from a square. Iterations are rowwise from top to bottom. In the left column of subfigures the clusters are computed by assigning the raster cells to the nearest centre. In the right column of subfigures the centres of the clusters are computed."}
interval <- 1 / 10
x <- y <- seq(from = interval / 2, to = 1, by = interval)
xy <- expand.grid(x = x, y = y)
set.seed(34526)
xsam <- runif(4)
ysam <- runif(4)
xysam <- data.frame(x = xsam, y = ysam)

i <- 1
repeat {
#compute distance matrix
  dx <- outer(X = xy$x, Y = xysam$x, FUN = "-")
  dy <- outer(X = xy$y, Y = xysam$y, FUN = "-")
  d <- sqrt(dx^2 + dy^2)

#cluster the gridnodes
  xy$cluster <- apply(X = d, MARGIN = 1, FUN = which.min)
  xy$cluster <- factor(xy$cluster)
if (i %in% c(1, 2, 8)) {
  print(ggplot() +
    geom_tile(data = xy, mapping = aes(x = x, y = y, fill = cluster, alpha = 0.5), colour = "black") +
    scale_fill_viridis_d() +
    geom_point(data = xysam, mapping = aes(x = x, y = y), shape = 16, size = 5) +
    scale_x_continuous(name = "", limits = c(0, 1), breaks = c(0.2, 0.4, 0.6, 0.8)) +
    scale_y_continuous(name = "", limits = c(0, 1), breaks = c(0.2, 0.4, 0.6, 0.8)) +
    coord_fixed() +
    theme(legend.position = "none"))
  }

#save sample of previous run to check convergence
  xysamcur <- xysam

#compute the centroids of clusters
  xysam$x <- tapply(xy$x, INDEX = as.factor(xy$cluster), FUN = mean)
  xysam$y <- tapply(xy$y, INDEX = as.factor(xy$cluster), FUN = mean)


if (i %in% c(1, 2, 9)) {
  print(ggplot() +
    geom_tile(data = xy, mapping = aes(x = x, y = y, fill = cluster, alpha = 0.5), colour = "black") +
    scale_fill_viridis_d() +
    geom_point(data = xysam, mapping = aes(x = x, y = y), shape = 16, size = 4) +
    scale_x_continuous(name = "", limits = c(0, 1), breaks = c(0.2, 0.4, 0.6, 0.8)) +
    scale_y_continuous(name = "", limits = c(0, 1), breaks = c(0.2, 0.4, 0.6, 0.8)) +
    coord_fixed() +
    theme(legend.position = "none"))
  }

#check convergence
  dxsam <- (xysam$x - xysamcur$x)^2
  dysam <- (xysam$y - xysamcur$y)^2
  dxysam <- sqrt(dxsam + dysam)
  sumdxysam <- sum(dxysam)
  if (sumdxysam < 1E-12) {
    break
    }

i <- i + 1
}
```

The same algorithm was used in Chapter \@ref(STSI) to construct compact geographical strata (shortly referred to as geostrata) for stratified random sampling. The clusters serve as strata. In stratified random sampling, one or more sampling units are selected randomly from each geostratum. However, for mapping purposes probability sampling is not required, so the random selection of a unit within each stratum is not needed. With random selection the spatial coverage is suboptimal. Here the centres of the final clusters (geostrata) are used as sampling points. This improves the spatial coverage compared to stratified *random* sampling.

In probability sampling we may want to have strata of equal area (clusters of equal size) so that the sampling design becomes self-weighting. For mapping equally sized clusters are not recommended as it may lead to samples with suboptimal spatial coverage.

```{block2, type = 'rmdnote'}
In Figure \@ref(fig:spatialcoveragesamplefromsquare) the clusters are of equal size but this is an artefact. Equally sized clusters are not guaranteed by the illustrated k-means algorithm. Clustering the raster cells of a square into four clusters is a very special case. In other cases the clusters computed with the k-means algorithm described above  might well have unequal size. In package **spcosa** also a different k-means algorithm is implemented, using swops, enforcing compact clusters  of equal size. 
```

Spatial coverage samples can be computed with package **spcosa** [@walvoort2010], using functions `stratify` and `spsample`, see code chunk below. Argument `nTry` of function ` stratify` specifies the number of initial stratifications in k-means clustering. Note that function `spsample` of package **spcosa** without optional argument `n` selects non-randomly one point in each cluster, being the centre. Figure \@ref(fig:spatcovVoorst) shows a spatial coverage sample of the same size as the regular grid in study area Voorst (Figure \@ref(fig:gridVoorst)). Note that the undersampled area in the centre of the study area is now covered by a sampling point.

```{r}
library(spcosa)
n <- 115
set.seed(314)
gridded(grdVoorst) <- ~ s1 + s2
mystrata <- spcosa::stratify(
  grdVoorst, nStrata = n, equalArea = FALSE, nTry = 10)
mysample <- spsample(mystrata)
mysample <- as(mysample, "data.frame")
```

```{r spatcovVoorst, echo = FALSE, out.width = '100%', fig.cap = "Spatial coverage sample from Voorst."}
mystrata <- as(mystrata,  "data.frame")
ggplot(mystrata) +
    geom_raster(mapping = aes(x = s1 / 1000, y = s2 / 1000, fill = factor(stratumId))) +
    scale_fill_viridis_d(name = "geostratum") +
    geom_point(data = mysample, mapping = aes(x = s1 / 1000, y = s2 / 1000), size = 1, colour = "orange") +
    coord_fixed() +
    scale_x_continuous(name = "Easting (km)") +
    scale_y_continuous(name = "Northing (km)") +
    theme(legend.position = "none") +  theme(panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_blank())
```

If the clusters need not be of equal size, we may also use function `kmeans` of the **stats** package, using the spatial coordinates as clustering variables. This requires less computing time, especially with large data sets.

```{r}
grdVoorst <- as(grdVoorst, "data.frame")
mystrata_kmeans <- kmeans(
  grdVoorst[, c("s1", "s2")], centers = n, iter.max = 10000, nstart = 10)
mysample_kmeans <- mystrata_kmeans$centers %>% data.frame(.)
```

When function `kmeans` is used to compute the spatial coverage sample, there is no guarantee that the computed centres of the clusters, used as sampling points, are inside the study area. In Figure \@ref(fig:kmeanscenters) there are eight such centres.

(ref:kmeanscenters) Centres of spatial clusters computed with function `kmeans`.

```{r kmeanscenters, echo = FALSE, out.width = '100%', fig.cap = "(ref:kmeanscenters)"}
gridded(grdVoorst) <- ~ s1 + s2
coordinates(mysample_kmeans) <- ~ s1 + s2
res <- over(mysample_kmeans, grdVoorst)
inside <- as.factor(!is.na(res$z))
levels(inside) <- (c("Outside", "Inside"))
grdVoorst <- as(grdVoorst, "data.frame")
mysample_kmeans <- as(mysample_kmeans, "data.frame")
ggplot() +
  geom_raster(data = grdVoorst, mapping = aes(x = s1 / 1000, y = s2 / 1000), fill = "grey") +
  geom_point(data = mysample_kmeans, mapping = aes(x = s1 / 1000, y = s2 / 1000, shape = inside), size = 1.5) +
  scale_shape_manual(values = c(8, 3), name = "") +
  scale_x_continuous(name = "Easting (km)") +
  scale_y_continuous(name = "Northing (km)") +
  theme_minimal() +
  coord_fixed() +
  theme(panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_blank())
```

This problem can easily be solved by selecting points inside the study area closest to the centres that are outside the study area. Function `rdist` of package **fields** is used to compute a matrix with distances between the centres outside the study area and the nodes of the discretisation grid. Then function `apply` is used with argument `FUN = which.min` to compute the discretisation nodes closest to the centres outside the study area. A similar procedure is implemented in function `spsample` of package **spcosa** when the centres of the clusters are selected as sampling points (so, when argument `n` of function `spsample` is not used).

```{r}
library(fields)
units_out <- which(inside == FALSE)
D <- fields::rdist(x1 = mysample_kmeans[units_out, ],
  x2 = grdVoorst[, c("s1", "s2")])
units_close <- apply(D, MARGIN = 1, FUN = which.min)
mysample_kmeans[units_out, ] <- grdVoorst[units_close, c("s1", "s2")]
```

#### Exercises {-}

2. In forestry and vegetation surveys square and circular plots are often used as sampling units, for instance, 2 m squares or circles with a diameter of 2 m. To study the relation between the vegetation and the soil, soil samples must be collected from the vegetation plots. Suppose we want to collect four soil samples from a square plot. Where would you locate the four sampling points, so that they optimally cover the plot?  
3. Suppose  we are also interested in the accuracy of the estimated plot means of the soil properties, not just the means. In that case the soil samples should not be bulked into a composite sample, but analysed separately. How would you select the sampling points in this case?  
4. For circular vegetation plots it is less clear where the sampling points with smallest MSSD (Equation \@ref(eq:MSSD)) are. Write an **R** script to compute a spatial coverage sample of five points from a circular plot discretised by the nodes of a fine square grid.  Use argument `equalArea = FALSE`. Check the size (number of raster cells) of the strata.  Repeat this for six sampling points.  
5. Consider the case of six strata. The strata are not of equal size. If the soil samples are bulked into a composite sample, the measurement on this single sample is a biased estimator of the plot mean. How can this bias be avoided?  

## Spatial infill sampling {#SpatialInfill}

If georeferenced data are available that can be used for mapping the study variable, but we need more data for mapping, it is attractive to account for these existing sampling units when selecting the additional units. The aim now is to fill in the empty spaces, i.e. the parts of the study area not covered by the existing sampling units. This is referred to as *spatial infill sampling*. Existing sampling units can easily be accommodated in the k-means algorithm, using them as fixed cluster centres\index{Fixed cluster centre}.

Figure \@ref(fig:spatialinfillEthiopia) shows a spatial infill sample for West-Amhara. A large set of legacy data on soil organic matter (SOM) in mass percentage (dag kg^-1^) is available, but these data come from strongly spatially clustered units along roads (the prior points in Figure \@ref(fig:spatialinfillEthiopia)). This is a nice example of a convenience sample. The legacy data are not ideal for mapping the SOM concentration throughout West-Amhara. Clearly, it is desirable to collect additional data in the off-road parts of the study area, with the exception of the northeastern part where we have already quite a few data not near the main roads. The legacy data are passed to function `stratify` of package **spcosa** with argument `priorPoints`. The object assigned to this argument must be of class `SpatialPoints` or  `SpatialPointsDataFrame`. This optional argument fixes these points as cluster centres. A spatial infill sample of 100 points is selected, taking into account these fixed points.

<!--Projection attributes of the file with legacy sample: crs <- st_crs("+proj=laea +ellps=WGS84 +lat_0=5 +lon_0=20 +no_defs")-->

```{r spatialinfillEthiopia, out.width="100%", fig.cap = "Spatial infill sample of 100 points from West-Amhara."}
gridded(grdAmhara) <- ~ s1 + s2
n <- 100
ntot <- n + nrow(sampleAmhara)
coordinates(sampleAmhara) <- ~ s1 + s2
proj4string(sampleAmhara) <- NA_character_
set.seed(314)
mystrata <- spcosa::stratify(grdAmhara, nStrata = ntot,
  priorPoints = sampleAmhara, nTry = 10)
mysample <- spsample(mystrata)
plot(mystrata, mysample)
```

In the output object of `spsample` both the prior and the new sampling points are included. The new points can be obtained as follows:

```{r}
units <- which(mysample@isPriorPoint == FALSE)
mysample <- as(mysample, "data.frame")
mysample_new <- mysample[units, ]
```

#### Exercises {-}
6. Write an **R** script to select a spatial infill sample of size 100 from study area Xuancheng in China. Use the iPSM sample in `data.frame` `sampleXuancheng` of package **sswr** as a legacy sample. To map the SOM concentration we want to measure the SOM concentration at 100 more sampling points. 
    +  Read the file `data/Elevation_Xuancheng.rds` with function `rast` of package **terra**, and use this file as a discretisation of the study area.
    +  For computational reasons there are far too many raster cells. That many cells are not needed to select a spatial infill sample. Subsample the raster file by selecting a square grid with a spacing of 900 m $\times$ 900 m. First convert the `SpatRaster` object to a `data.frame`, and then change it to a `SpatialPixelsDataFrame` using function `gridded`. Then use function `spsample` with argument `type = "regular"`.
    +  Select a spatial infill sample using functions `stratify` and `sample` of package **spcosa**.  
  
```{r, echo = FALSE}
rm(list = ls())
```

