# Covariate space coverage sampling {#kmeans}

Regular grid sampling and spatial coverage sampling are pure spatial sampling designs. Covariates possibly related to the study variable are not accounted for in selecting sampling units. This can be suboptimal when the study variable is related to covariates of which maps are available, think for instance of remote sensing imagery or digital elevation models related to soil properties. Maps of these covariate can be used in mapping the study variable by, for instance, a multiple linear regression model or a random forest. This chapter describes a simple, straightforward method for selecting sampling units on the basis of the covariate values of the grid cells. This is a pure feature (attribute) space design.

The simplest option for covariate space coverage sampling\index{Covariate space coverage sampling} (CSC sampling) is to cluster the grid cells by the $k$-means clustering algorithm in multivariate covariate space. Similar to spatial coverage sampling (Chapter \@ref(SpatialCoverage)) the mean squared shortest distance (MSSD) is minimised, but now the distance is not measured in geographical space but in a $p$-dimensional space spanned by the $p$ covariates (think of it as a multidimensional scatter plot with the covariates along the axes). The covariates are centered and scaled, so that their means become 0 and standard deviations become 1. This is needed because, contrary to the spatial coordinates used as clustering variables in spatial coverage sampling, the dimensions of the covariates used as clustering variables generally differ, and the ranges of the covariates in the population can differ greatly. In the clustering of the grid cells the mean squared shortest *scaled* distance (MSS*S*D)^[The name 'scaled distance' can be confusing. Not the geographic distances are scaled, rather, the distances are computed in a space spanned by the scaled covariates] is minimised.

In the next code chunk a CSC sample of size 20 is selected from Eastern Amazonia. To speed up the computations a subgrid with a spacing of 5000 m is selected, using function `spsample` of package **sp**. All five quantitative covariates, SWIR2, Terra_PP, Prec_dm, Elevation and Clay, are used as covariates. To select twenty points, twenty clusters are constructed using function `kmeans` of the **stats** package [@R2020]. The number of clusters is specified with argument `centers`. Note that the number of clusters is not based, as usual in cluster analysis, on the expected number of subregions with a high density of points in the multivariate distribution, but rather on the number of observation points to be sampled. The $k$-means clustering algorithm is a deterministic algorithm, i.e. the final optimised clustering is fully determined by the initial clustering. This final clustering can be suboptimal, i.e. the minimised MSSSD value is somewhat larger than the global minimum. Therefore the clustering should be repeated many times, every time starting with a different random initial clustering. The number of repeats is specified with argument `nstart`.

```{r}
load("data/Amazonia_5km.RData")
covs <- names(gridAmazonia)[c(4,5,6,7,8)]
n <- 20
set.seed(314)
myclusters <- kmeans(
  scale(gridAmazonia[,covs]), centers=n, iter.max=10000, nstart=100)
gridAmazonia$cluster <- myclusters$cluster
```

Grid cells with the shortest scaled Euclidean distance in covariate-space to the centers of the clusters are selected as the sampling points. To this end first a matrix with the distances of all the grid cells to the cluster centers is computed with function `rdist` of package **fields** [@fields]. The grid cells closest to the centers are computed with function `apply`, using argument `FUN=which.min`. 

```{r}
library(fields)
D <- rdist(x1=myclusters$centers, x2=scale(gridAmazonia[,covs]))
units <- apply(D, MARGIN=1, FUN=which.min)
myCSCsample <- gridAmazonia[units,]
```

Figure \@ref(fig:CSCsample) shows the clustering of the grid cells and the grid cells closest in covariate space to the centers,  used as the selected sample. In Figure \@ref(fig:CSCsampleinscatter) the selected sample is plotted in scatter diagrams of some pairs of covariates. In the scatter diagrams some sampling points are clearly clustered. However, this is misleading, as actually we must look in five-dimensional space to see whether the points are clustered. Two points with a large separation distance in a five-dimensional space can look quite close when these two points are projected on a two-dimensional plane, see also Exercise 2 of this chapter.

The next code chunk shows how the MSSSD of the selected sample can be computed.

```{r}
popmeans <- apply(gridAmazonia[,covs], MARGIN=2, FUN=mean)
popsds <- apply(gridAmazonia[,covs], MARGIN=2, FUN=sd)
D <- rdist(x1=scale(myCSCsample[,covs], center=popmeans, scale=popsds),
           x2=scale(gridAmazonia[,covs]))
dmin <- apply(D, MARGIN=2, min)
MSSSD <- mean(dmin^2)
```

Note that to center and scale the covariate values in the CSC sample, the population means and population standard deviations are used, as specified with arguments `center` and `scale` of function `scale`. If these means and standard deviations are unspecified, the *sample* means and *sample* standard deviations are used, resulting in an incorrect value of the minimised MSSSD value. The MSSSD of the selected sample equals `r as.character(round(MSSSD,3))`.

```{r CSCsample, echo=FALSE, out.width="100%", fig.cap="Covariate space coverage sample of twenty points obtained with $k$-means, from Eastern Amazonia using five covariates in clustering. The map shows the spatial distribution of the twenty clusters."}
ggplot(gridAmazonia) +
  geom_raster(mapping=aes(x=x1/1000, y=x2/1000, fill=as.character(cluster))) +
  geom_point(data=myCSCsample,mapping=aes(x=x1/1000,y=x2/1000), size=2, colour="red") +
  scale_fill_viridis_d(name="Cluster") +
  scale_x_continuous(name="Easting (km)") +
  scale_y_continuous(name="Northing (km)") +
  coord_fixed() +
  theme(legend.position="none")
```

```{r CSCsampleinscatter, echo=FALSE, out.width='100%', fig.asp=0.5, fig.cap="Covariate space coverage sample, obtained with $k$-means, plotted in scatter diagrams of pairs of covariates, coloured by cluster."}
plt1 <- ggplot(gridAmazonia) +
  geom_point(mapping=aes(x=SWIR2, y=Prec_dm, colour=as.character(cluster)), alpha=0.5) +
  scale_colour_viridis_d() +
  geom_point(data=myCSCsample, mapping=aes(x=SWIR2, y=Prec_dm), size=1.5, colour="red") +
  scale_x_continuous(name="SWIR2") +
  scale_y_continuous(name="Precipitation dryest month") +
  theme(legend.position="none") +
  theme(axis.title.x=element_text(size=14),
        axis.title.y=element_text(size=14))

plt2 <- ggplot(gridAmazonia) +
  geom_point(mapping=aes(x=Terra_PP, y=Elevation, colour=as.character(cluster)), alpha=0.5) +
  scale_colour_viridis_d() +
  geom_point(data=myCSCsample, mapping=aes(x=Terra_PP, y=Elevation), size=1.5, colour="red") +
  scale_x_continuous(name="Terra_PP") +
  scale_y_continuous(name="Elevation") +
  theme(legend.position="none") +
  theme(axis.title.x=element_text(size=14),
        axis.title.y=element_text(size=14))
grid.arrange(plt1, plt2, nrow=1)
```

Instead of function `kmeans` we may use function `kmeanspp` of package **LICORS** [@Goerg2013]. This function is an implementation of the  $k$-means++ algorithm [@Arthur2007]. This algorithm consists of two parts, namely the selection of an optimised initial sample, followed by the standard $k$-means. The algorithm is as follows:  

1. Select one unit (raster cell) at random.  
2. For each unit $j$, compute $d_{ij}$, the distance in standardised covariate-space between $j$ and the nearest unit $i$ that has already been selected.  
3. Choose one new raster cell at random as a new sampling unit with probabilities proportional to $d^2_{ij}$, and add the selected raster cell to the set of selected cells.    
4. Repeat Steps 2 and 3 until $n$ centers have been selected.  
5. Now that the initial centers have been selected, proceed using standard $k$-means.  

```{r, eval=FALSE}
library(LICORS)
myclusters <- kmeanspp(
  scale(gridAmazonia[,covs]), k=n, iter.max=10000, nstart=30)
```

Due to the improved initial centers, the risk of ending in a local minimum is reduced. The $k$-means++ algorithm\index{\emph{k}-means++ algorithm} is of most interest for small sample sizes. For large sample sizes the extra time needed for computing the initial centers can become substantial and may not outweigh the larger number of starts that can be afforded with the usual $k$-means algorithm for the same computing time.

## Covariate space infill sampling

If we have legacy data that can be used to fit a model for mapping, it is more efficient to select an infill sample, similar to spatial infill sampling explained in Section \@ref(SpatialInfill). The only difference with spatial infill sampling is that the legacy data are now plotted in the space spanned by the covariates, and the empty regions we would like to fill in are now the undersampled regions in this covariate space. The legacy sample units serve as fixed cluster centers, they cannot move through the covariate space during the optimisation of the infill sample. In the next code chunk a function is defined for covariate space infill sampling.

```{r}
CSIS <- function (fixed,nsup,nstarts,mygrd) {
  nfix <- nrow(fixed)
  p <- ncol(mygrd)
  units <- fixed$units
  mygrd_minfx <- mygrd[-units,]
  MSSSD_cur <- NA
  for (s in 1:nstarts) {
    units <- sample.int(nrow(mygrd_minfx),nsup)
    centers_sup <- mygrd_minfx[units,]
    centers <- rbind(fixed[,names(mygrd)],centers_sup)
    repeat {
      D <- rdist(x1=centers, x2=mygrd)
      clusters <- apply(X=D, MARGIN=2, FUN=which.min) %>% as.factor(.)
      centers_cur <- centers
      for (i in 1:p) {
        centers[,i] <- tapply(mygrd[,i], INDEX=clusters, FUN=mean)
      }
      #restore fixed centers
      centers[1:nfix,] <- centers_cur[1:nfix,]
      #check convergence
      sumd <- diag(rdist(x1=centers, x2=centers_cur)) %>% sum(.)
      if (sumd < 1E-12) {
        D <- rdist(x1=centers, x2=mygrd)
        Dmin <- apply(X=D, MARGIN=2, FUN=min)
        MSSSD <- mean(Dmin^2)
        if (s==1 | MSSSD < MSSSD_cur) {
          centers_best <- centers
          clusters_best <- clusters
          MSSSD_cur <- MSSSD
        }
        break
      }
    }
  }
  list(centers=centers_best, clusters=clusters_best)
}
```

The function is used to select an infill sample of fifteen units from Eastern Amazonia. A legacy sample of five units is randomly selected.

```{r, eval=FALSE}
set.seed(314)
units <- sample.int(nrow(gridAmazonia),5)
fixed <- data.frame(units, scale(gridAmazonia[,covs])[units,])
mygrd <- data.frame(scale(gridAmazonia[,covs]))
res <- CSIS(fixed=fixed, nsup=15, nstarts=10, mygrd=mygrd)
```

```{r, eval=FALSE, echo=FALSE}
saveRDS(res, file="results/CSIS_Amazonia.rds")
```

```{r, echo=FALSE}
res <- readRDS(file="results/CSIS_Amazonia.rds")
```

```{r CSCIS, echo=FALSE, out.width="100%", fig.cap="Covariate space infill sample, obtained with modified $k$-means from Eastern Amazonia in a map of SWIR2. The dots represent the fixed centers (legacy sample, five units), the triangles the infill sample (fifteen units)."}
gridAmazonia$cluster <- res$clusters
D <- rdist(x1=res$centers, x2=scale(gridAmazonia[,covs]))
units <- apply(D, MARGIN=1, FUN=which.min)
myCSIsample <- gridAmazonia[units,c("x1","x2",covs)]
myCSIsample$free <- c(rep("FALSE",5), rep("TRUE",15))

ggplot(data=gridAmazonia) +
  geom_raster(mapping=aes(x=x1/1000, y=x2/1000, fill=as.character(cluster))) +
#  geom_raster(mapping=aes(x=x1/1000, y=x2/1000, fill=SWIR2))+  
  geom_point(data=myCSIsample, mapping=aes(x=x1/1000, y=x2/1000, shape=free), colour="red", size=2) +
  scale_x_continuous(name="Easting (km)") +
  scale_y_continuous(name="Northing (km)") + 
  scale_fill_viridis_d(name="SWIR2") +
  coord_fixed() +
  theme(legend.position="none")
```

```{r CSInfillsampleinscatter, echo=FALSE, out.width='100%', fig.asp=0.5, fig.cap="Covariate space infill sample, obtained with modified $k$-means, plotted in scatter diagrams of pairs of covariates. The dots represent the fixed centers (legacy sample), the triangles the infill sample."}
plt1 <- ggplot(gridAmazonia) +
  geom_point(mapping=aes(x=SWIR2, y=Prec_dm, colour=as.character(cluster)), alpha=0.5) +
  scale_colour_viridis_d() +
  geom_point(data=myCSIsample, mapping=aes(x=SWIR2, y=Prec_dm, shape=free), size=1.5, colour="red") +
  scale_x_continuous(name="SWIR2") +
  scale_y_continuous(name="Precipitation dryest month") +
  theme(legend.position="none") +
  theme(axis.title.x=element_text(size=14),
        axis.title.y=element_text(size=14))

plt2 <- ggplot(gridAmazonia) +
  geom_point(mapping=aes(x=Terra_PP, y=Elevation, colour=as.character(cluster)), alpha=0.5) +
  scale_colour_viridis_d() +
  geom_point(data=myCSIsample, mapping=aes(x=Terra_PP, y=Elevation, shape=free), size=1.5, colour="red") +
  scale_x_continuous(name="Terra_PP") +
  scale_y_continuous(name="Elevation") +
  theme(legend.position="none") +
  theme(axis.title.x=element_text(size=14),
        axis.title.y=element_text(size=14))
grid.arrange(plt1, plt2, nrow=1)
```

## Performance of covariate space coverage sampling in random forest prediction

Covariate space coverage sampling can be a good candidate sampling design if we have multiple maps of covariates and we do not want to rely on a linear relation between the study variable and the covariates. In this situation we may consider mapping with machine learning algorithms\index{Machine learning technique}, such as neural networks and random forests\index{Random forest}.

I used the Eastern Amazonia data set to evaluate CSC sampling for mapping the aboveground biomass (AGB). The five covariates are used as predictors in random forest modelling. The calibrated models are used to predict AGB at the units of a validation sample of size 25000, selected by simple random sampling without replacement from the 1 km $\times$ 1 km grid, excluding the cells of the 10 km $\times$ 10 km grid from which the calibration samples are selected. The predicted AGB values at the validation units are compared with the true AGB values, and the prediction errors computed. The sample mean of the (squared) prediction error is a design-unbiased estimator of the population mean (squared) error, i.e. the mean of the (squared) errors at all population units (excluding the units of the 10 km $\times$ 10 km grid), see Chapter \@ref(Validation). Besides, the Nash-Sutcliffe model efficiency coefficient\index{Model efficiency coefficient} (MEC) is estimated.

Three sample sizes are used, $n=$ 25, 50, 100. Of each sample size 500 CSC samples are selected using the $k$-means algorithm, leading to 1500 CSC samples in total.  The number of starts are 500, 350 and 200 for $n=$ 25, 50, 100, respectively. With these number of starts the computing time was about equal to conditioned Latin hypercube sampling, see next chapter. Each sample is used to calibrate a random forest model. Simple random sampling (SI) is used as a reference strategy. The results are described in detail in the next chapter. In short: for $n=25$ and $50$ CSC performs on average somewhat better than SI, for $n=100$ they perform about equal. Most striking is the smaller spread in the map quality indices with CSC as compared to SI.  

In Figure \@ref(fig:RelationMSSSDRMSE) the RMSE is plotted against the minimised MSSSD, both for the $3 \times 500$ CSC samples, and for the $3 \times 500$ simple random samples. The vertical strings of points in the scatter plots are the points for the CSC samples. For all three sample sizes the minimised MSSSD values of the CSC samples is substantially smaller than those of the SI samples. For $n=25$ and $50$ the slope of the fitted line is positive: the smaller the MSSSD the smaller the expected RMSE. For $n=100$ the estimated slope parameter is slightly negative. So for this sample size we do not profit from spreading the sampling units in multivariate covariate space through $k$-means clustering.

```{r RelationMSSSDRMSE, echo=FALSE, fig.asp=0.7, fig.cap="Relation between minimisation criterion MSSSD and the root mean squared error (RMSE) of random forest predictions of aboveground biomass in Eastern Amazonia for three sample sizes. Solid lines are the fitted simple linear regression models for RMSE with MSSSD as a predictor."}
load("results/CSCversusCLH_Amazonia_n25.RData")
df.25 <- data.frame(RMSE=c(RMSE.SI,RMSE.CSC), MSSSD=c(MSSSD.SI,MSSSD.CSC), Design=rep(c("SI","CSC"), each=500),Samplesize=25)

load("results/CSCversusCLH_Amazonia_n50.RData")
df.50 <- data.frame(RMSE=c(RMSE.SI,RMSE.CSC), MSSSD=c(MSSSD.SI,MSSSD.CSC), Design=rep(c("SI","CSC"), each=500),Samplesize=50)

load("results/CSCversusCLH_Amazonia_n100.RData")
df.100 <- data.frame(RMSE=c(RMSE.SI,RMSE.CSC), MSSSD=c(MSSSD.SI,MSSSD.CSC), Design=rep(c("SI","CSC"), each=500),Samplesize=100)

df <- rbind(df.25, df.50, df.100)
df$Samplesize <- as.factor(df$Samplesize)
RMSE.lm <- lm(RMSE~MSSSD*Samplesize, data=df)
ggplot(df, mapping=aes(x=MSSSD, y=RMSE, shape=Samplesize))+
  geom_point(alpha=0.5) +
  scale_shape_manual(values=c(1,0,2), name="Sample size") +
  geom_line(mapping=aes(x=MSSSD, y=RMSE.lm$fitted))
```

#### Exercises {-}

1. Write an **R** script to select a covariate space coverage (CSC) sample of size 20,  using the covariates cti and ndvi in $k$-means clustering of the grid cells. Plot the clusters and the sample in a scatter diagram of cti against ndvi.  
2. Repeat this, but now with covariates cti, ndvi *and* elevation_m as covariates. Plot the clusters and the sample again in a scatter diagram of cti against ndvi. Explain that in this case a sampling location (cluster center) can be quite close to another sampling location, and that a cluster in a 2D-scatter diagram may also contain units of a different cluster.   
  
```{r, echo=FALSE}
rm(list=ls())
```

