# Spatial response surface sampling {#SpatialResponseSurface}

As with conditioned Latin hypercube sampling (cLHS), spatial response surface sampling\index{Spatial response surface sampling} is an experimental design adapted for spatial surveys. Experimental response surface designs aim at finding an optimum combination of the response within specified ranges of the factors. There are many types of response surface designs, see @myers2002.  With response surface sampling one assumes that some type of low order (linear or quadratic) regression model can be used to accurately approximate the relationship between the study variable and the covariates.  A commonly used design is the central composite design\index{Central composite design}. The data produced as a result of this design are used to fit a curved, quadratic surface, i.e. a multiple linear regression model with quadratic terms.

The response surface sampling approach is an example of a model-based sampling design. From that viewpoint I should have described this sampling design in the later part of this book dealing with model-based sampling. Sampling units are selected to implicitly optimise the estimation of the linear or quadratic regression model. However, this optimisation is done under one or more spatial constraints. Unconstrained optimisation of the sampling design under the linear regression model will not prevent the units from spatial clustering, see Figure \@ref(fig:twosamples)(b). The assumption of independent data might be violated when the sampling units are spatially clustered.  For that reason the response surface sampling design selects samples with good spatial coverage, so that the design becomes robust against violation of the independence assumption.

@lesch95 adapted the response surface methodology so that it can be applied in observational studies. Several problems needed to be tackled. First, when multiple covariates are used, the covariates must be decorrelated. Second, when sampling units are spatially clustered, the assumption in linear regression modelling of spatially uncorrelated model residuals can be violated. To address these two problems @lesch95 proposed the following procedure; see also @lesch2005:   

1. Transform the covariate matrix into a scaled, centered, de-correlated matrix by principal components analysis (PCA).   
2. Choose the response surface design type.   
3. Select candidate sampling units based on the distance from the design points in PC-space and then select multiple units per design point.   
4. Select the combinatin of candidate sampling units with the highest value for a criterion that quantifies how uniform the sample is spread across the study area.   

This design has been applied, among others, for mapping soil salinity (ECe), using electromagnetic induction (EM) measurements and surface array conductivity measurements as predictors in multiple linear regression models. For applications, see @corwin2005, @lesch2005, @fitzgerald2006, @Corwin2010 and @Fitzgerald2010.

Spatial response surface sampling  is illustrated with the electromagnetic induction (EM) measurements (mS/m) of the apparent electrical conductivity on the 80 ha Cotton Research Farm in Uzbekistan (Section \@ref(Datasets)). The EM measurements in vertical dipole mode with transmitter at 1 m and 0.5 m from the receiver, are on transects covering the Cotton Research Farm (Figure \@ref(fig:EMdataUzbekistan)). As a first step the natural logarithm of the two EM measurements are interpolated by ordinary kriging to a fine grid (Figure \@ref(fig:EMdataUzbekistan2)). These ordinary kriging predictions of lnEM are used as covariates in response surface sampling (Figure \@ref(fig:EMdataUzbekistan)). The two covariates are strongly correlated, $r=0.73$, as would be expected since they are of the same variable at two depths.

```{r EMdataUzbekistan, echo=FALSE, out.width='100%', fig.cap="Natural logarithm of EM38 measurements on the Cotton Research Farm in Uzbekistan (with transmitter at 1 m and 0.5 m from receiver)."}
dat <- read.csv(file="data/TransectsData_EM_CRF.csv")
dat$lnEMv1m <- log(dat$EMv1mt)
dat$lnEMv0.5m <- log(dat$EMv05mt)
dat <- na.omit(dat)
coordinates(dat) <- ~ x * y
proj4string(dat) <- CRS("+proj=longlat +datum=WGS84") #=  +init=epsg:4326
dat <- spTransform(dat, CRS = CRS("+proj=tmerc +lat_0=0.000000000 +lon_0=63.000000000 +k=1.000000 +x_0=500000.000 +y_0=0.000 +ellps=krass +units=m"))
dat <- as(dat,"data.frame")
names(dat)[c(9,10)] <- c("lnEM100cm","lnEM50cm")
df <- dat %>% pivot_longer(cols=c("lnEM100cm","lnEM50cm"))

ggplot(data = df) +
  geom_point(mapping = aes(x = x/1000, y = y/1000, colour = value), size=0.3) +
  scale_colour_viridis_c(name = "lnEM") +
  scale_x_continuous(name="Easting (km)",limits=c(305.5,308)) +
  scale_y_continuous(name="Northing (km)",limits=c(4602,4603.5)) +
  coord_fixed() +
  facet_grid(~ name)
```



```{r EMdataUzbekistan2, echo=FALSE, out.width='100%', fig.cap="Interpolated surfaces of natural logarithm of EM38 measurements on the Cotton Research Farm in Uzbekistan, used as covariates in spatial response surface sampling."}
load(file="data/CottonResearchFarm.RData")
df <- EM_CRF %>% pivot_longer(cols=c("lnEM100cm","lnEM50cm"))

ggplot(data = df) +
  geom_raster(mapping = aes(x = x1/1000, y = x2/1000, fill = value)) +
  scale_fill_viridis_c(name = "lnEM") +
  scale_x_continuous(name="Easting (km)",limits=c(305.5,308)) +
  scale_y_continuous(name="Northing (km)",limits=c(4602,4603.5)) +
  coord_fixed() +
  facet_grid(~ name)
```

Function `prcomp` of the **stats** package [@R2020] is used to compute the principal component scores\index{Principal component score} for all units in the population (grid cells). The two covariates are centered and scaled, i.e. we compute standardised principal components. 

```{r}
grdsub <- subset(EM_CRF, select=c(lnEM100cm,lnEM50cm))
pc <- prcomp(grdsub, center=TRUE, scale=TRUE)
```

The means of the two principal component scores are 0, their standard deviations are `r round(pc$sdev[1],3)` and `r round(pc$sdev[2],3)`. The principal component scores are then scaled so that their standard deviations become 1, i.e. they will have the same weight in the following steps. 

```{r}
EM_CRF$PC1 <- pc$x[,1]*1/pc$sdev[1]
EM_CRF$PC2 <- pc$x[,2]*1/pc$sdev[2]
```

Figure \@ref(fig:candidatelocations) shows a scatter plot of the two standardised principal component scores of all `r nrow(EM_CRF)` grid cells.

Function `ccd` of package **rsm** ([@Lenth2009] is now used to generate a central composite response surface design\index{Central composite response surface design} (CCRSD). Argument `basis` specifies the number of factors, which is two in our case; argument `n0` is the number of center points, and argument `alpha` determines the position of the "star points" (explained hereafter).

```{r}
library(rsm)
set.seed(314)
print(ccdesign <- ccd(basis=2, n0=1, alpha="rotatable"))
```

The experiment consists of two blocks, each of five experimental units. Block 1, the so-called cube block, consists of one center point and four cube points\index{Cube point}. In the experimental unit represented by the center point both factors have levels in the center of the experimental range. In the experimental units represented by the cube points the levels of both factors is either -1 or +1 unit in the design space. Block 2, referred to as the star block, consists of one center point and four star points\index{Star point}. With `alpha = rotatable` the  start points are on the circle circumscribing the square (Figure \@ref(fig:ccdesign)).

```{r ccdesign, echo=FALSE, fig.cap="Rotatable central composite response surface design for two factors."}
cube <- data.frame(x1=c(-1,1,-1,1,0),x2=c(-1,-1,1,1,0))
star <- data.frame(x1=c(-sqrt(2),sqrt(2),0,0,0), x2=c(0,0,-sqrt(2),sqrt(2),0))
df <- rbind(cube,star)
df$block <- rep(c("cube","star"),each=5)
ggplot(df) +
  geom_point(mapping=aes(x=x1,y=x2,shape=block),size=2) +
  scale_shape_manual(values=c(0, 8)) +
  geom_path(data=data.frame(x=c(-1,-1,1,1,-1),y=c(-1,1,1,-1,-1)),aes(x=x,y=y),lty=2) +
  geom_circle(aes(x0=0,y0=0,r=sqrt(2)),lty=2) +
  coord_fixed()
```

To adapt this design for an observational study, we drop one of the center points (0,0).
 
```{r}
ccd_df <- data.frame(x1=ccdesign$x1, x2=ccdesign$x2)
ccd_df <- ccd_df[-6,]
```

The points in the CCRSD design are adjusted so that a large proportion $p$ of the bivariate standardised principal component scores of the population units are within a circle of a given radius. This radius is computed as a sample quantile of the empirical distribution of the distances of the points in the scatter to the center. For $p$ I chose 0.7.

```{r}
d <- sqrt(EM_CRF$PC1^2+EM_CRF$PC2^2)
radius <- quantile(d, p=0.7)
print(radius)
ccd_df$x1 <- ccd_df$x1*radius
ccd_df$x2 <- ccd_df$x2*radius
```

The next step is to select for each design point several candidate sampling units. For each of the nine design points\index{Design point} 0.5% of the grid cells are selected that are closest to that design point. This results in $9 \times 8$ candidate sampling points.

```{r}
EM_CRF$unit <- 1:nrow(EM_CRF)
candi_all <- NULL
for (i in 1:nrow(ccd_df)) {
    d2dpnt <- sqrt((EM_CRF$PC1-ccd_df$x1[i])^2+
                     (EM_CRF$PC2-ccd_df$x2[i])^2)
    q <- quantile(d2dpnt, p=0.005)
    units_close <- which(d2dpnt < q)
    candi <- EM_CRF[units_close,c("unit","x1","x2","PC1","PC2")]
    candi$dpnt <- i
    candi_all <- rbind(candi_all,candi)
}
```

Figure \@ref(fig:candidatelocations) shows the nine clusters of candidate sampling points around the design points. Note that the location of the candidate points associated with the design points with coordinates  (-1.51,-1.51), (0, -2.13), (1.51,-1.51) and (2.13,0) are all far inside the circle that passes through the design points. So for the optimised sample the value of one of the principal component scores will be less extreme than the ideal values according to the CCRSD design.   

```{r candidatelocations, echo=FALSE, fig.cap="Clusters of units (red points) around the design points (triangles) of a central composite design (two covariates), serving as candidate sampling points."}
notcandidates <- EM_CRF[-candi_all$unit,]
ggplot(notcandidates) +
  geom_point(mapping=aes(x=PC1,y=PC2),alpha=0.5) +
  geom_point(data=candi_all,mapping=aes(x=PC1,y=PC2),colour="red") +
  geom_point(data=ccd_df,aes(x=x1,y=x2),shape=2) +
  scale_x_continuous(name="PC1",limits=c(-4,3)) +
  scale_y_continuous(name="PC2",limits=c(-3,3)) +
  coord_fixed()
```
Figure \@ref(fig:candidatesingeospace) shows that in geographical space for most design points there are multiple clusters of candidate units. For instance, for design point 9 there are three clusters of candidate sampling units. Due to this there is scope to optimise the sample computationally. 

```{r candidatesingeospace, echo=FALSE, fig.cap="Candidate sampling points plotted in geographical space on the first standardised principal component."}
library(ggpubr)
ggplot(EM_CRF) +
  geom_raster(aes(x=x1/1000,y=x2/1000,fill=PC1)) +
  geom_point(data=candi_all,mapping=aes(x=x1/1000,y=x2/1000,shape=as.factor(dpnt)),size=2) +
  scale_shape_manual(values=c(0,1,2,3,4,5,6,7,8),name="design point") +
  scale_fill_viridis_c(name="PC1") + 
  scale_x_continuous(name="Easting (km)") +
  scale_y_continuous(name="Northing (km)") +
  coord_fixed()
```

As a first step an initial subsample from the candidate sampling units is selected by stratified simple random sampling, using the factor `dpnt` as strata. Function `strata` of package **sampling** is used for stratified random sampling [@Tille2016].

```{r}
library(sampling)
set.seed(314)
units_stsi <- sampling::strata(
  candi_all, stratanames="dpnt", size=rep(1,9))
mysample0 <- getdata(candi_all, units_stsi)
mysample0 <- mysample0[,c("unit","x1","x2","PC1","PC2","dpnt")]
```

The locations of the nine sampling points are now optimised by minimising a criterion. Two minimisation criteria are implemented, a geometric criterion and a model-based criterion. 

In the geometric criterion (as proposed by @lesch2005) for each sampling point the shortest distance to the other points is computed, and the logarithm of this shortest distance is computed. The minimisation criterion is the negative of the sample mean of the logarithm of shortest distance. 

The model-based minimisation criterion is the average correlation of the sampling points. This criterion requires as input the parameters of a residual correlogram (see Section \@ref(IntroKED)). I assume an exponential  correlogram without nugget, so that the only parameter to be chosen is the distance parameter $\phi$. Three times $\phi$ is referred to as the effective range\index{Effective range} of the exponential covariance function. The correlation of two random variables at this distance is 0.05. A penalty term is added to these criteria, equal to the average distance, in the space spanned by the principal component scores, of the sampling points to the associated design points, multiplied by a weight. With weights $> 0$ sampling points close to the design points are preferred over more distant points.

A function is defined for computing the minimisation criterion.  Given a chosen value for $\phi$, the $9 \times 9$ distance matrix of the sampling points can be converted into a correlation matrix, using function `variogramLine` of package **gstat** [@peb04]. Argument `weight` is an optional argument with default value 0.

```{r}
getCriterion <- function(mysample, dpnt, weight=0, phi=NULL){
  D2dpnt <- sqrt((mysample$PC1-dpnt$x1)^2+
                   (mysample$PC2-dpnt$x2)^2)
  D <- as.matrix(dist(mysample[,c("x1","x2")]))
  if (!is.null(phi)) {
    vgmodel <- vgm(model="Exp", psill=1, range=phi)
    C <- variogramLine(vgmodel, dist_vector=D, covariance=TRUE)
    criterion_cur <- mean(C)+mean(D2dpnt)*weight
  } else {
    diag(D) <- NA
    logdmin <- apply(D, MARGIN=1, FUN=min, na.rm=TRUE) %>% log(.)
    criterion_cur <- mean(-logdmin)+mean(D2dpnt)*weight
  }
}
```

The function `getCriterion` is used to compute the geometric criterion for the initial sample.

```{r}
criterion_geo <- getCriterion(mysample=mysample0, dpnt=ccd_df)
```

The initial value of the geometric criterion  is `r round(criterion_geo,3)`. The initial value for the model-based criterion is computed with an effective range of 50 m. Note that it does not make sense to make this range smaller than the size of the grid cells, which is 25 m in our case. For smaller ranges the correlation matrix is for any sample a matrix with zeroes. If the effective range is smaller than the smallest distance between two points in a cluster, the mean correlation is equal for all samples. 

```{r}
phi <- 50
criterion_mb <- getCriterion(
  mysample=mysample0, dpnt=ccd_df, phi=phi)
```

The initial value of the model-based criterion is `r round(criterion_mb,3)`.

The objective function defining the minimisation criterion is minimised with simulated annealing\index{Simulated annealing} (@Kirkpatrick1983, @Aarts1987). One sampling point is randomly selected, and replaced by another candidate sampling point from the same cluster. If the criterion of the new sample is smaller than that of the current sample, the new sample is accepted, if it is larger it is accepted with a probability that is a function of the change in the criterion (the larger the increase, the smaller the acceptance probability) and of an annealing parameter named the temperature (the higher the temperature, the larger the probability of accepting a new, poorer sample, given an increase of the criterion). See Section \@ref(SSA) for a more detailed introduction to simulated annealing. The annealing functions for spatial response surface sampling are defined in **R** script
[AnnealingFunctions4SpatialResponseSurfaceSampling.R](https://github.com/DickBrus/SpatialSamplingwithR/tree/master/Rscripts/AnnealingFunctions4SpatialResponseSurfaceSampling.R)^[https://github.com/DickBrus/SpatialSamplingwithR/tree/master/Rscripts/AnnealingFunctions4SpatialResponseSurfaceSampling.R].

```{r, eval=FALSE}
source('Rscripts/AnnealingFunctions4SpatialResponseSurfaceSampling.R')
set.seed(314)
mySRSsample <- anneal(
  mysample=mysample0, candidates=candi_all, dpnt=ccd_df,
  phi=50, T_ini=1, coolingRate=0.9,
  maxPermuted=25*nrow(mysample0), maxNoChange=20,
  verbose=TRUE)
```

```{r, eval=FALSE, echo=FALSE}
save(mySRSsample, file="results/SpatialResponseSurfaceSample_CRF_geo.RData")
#save(mySRSsample, file="results/SpatialResponseSurfaceSample_CRF_mb.RData")
```

Figure \@ref(fig:CCRSDinPCspace) shows the optimised CCRSD samples, obtained with the geometric and model-based criterion, plotted together with the design points. The two optimised samples are very similar.

```{r CCRSDinPCspace, echo=FALSE, out.width='100%', fig.cap="Principal component scores of CCRSD sample (triangles), optimised with the geometric and model-based criterion. Dots: design points."}
load(file="results/SpatialResponseSurfaceSample_CRF_geo.RData")
mysample_geo<-mySRSsample$mysample
critmin_geo <- tail(mySRSsample$trace,1)

load(file="results/SpatialResponseSurfaceSample_CRF_mb.RData")
mysample_mb <- mySRSsample$mysample
critmin_mb <- tail(mySRSsample$trace,1)

mysamples <- rbind(mysample_geo, mysample_mb)
mysamples$design <- rep(c("geometric","model-based"), each=9)

ggplot(mysamples) +
 geom_point(aes(x=PC1,y=PC2), shape=2, size=3) +
 geom_point(data=ccd_df, mapping=aes(x=x1,y=x2), size=3) +
 scale_x_continuous(name="PC1") +
 scale_y_continuous(name="PC2") +
 facet_wrap(~ design) +
 coord_fixed()
```

Figure \@ref(fig:CCRSDSample) shows the two optimised CCRSD samples plotted in geographical space on the first standardised principal component scores.

```{r CCRSDSample, echo=FALSE, fig.cap="CCRSD sample from the Cotton Research Farm in Uzbekistan, optmized with the geometric and model-based criterion."}

ggplot(mysamples) +
  geom_raster(data=EM_CRF, mapping=aes(x=x1/1000,y=x2/1000,fill=PC1)) +
  geom_point(data=mysamples, mapping=aes(x=x1/1000, y=x2/1000), size=2) +
  scale_fill_viridis_c(name="PC1") + 
  scale_x_continuous(name="Easting (km)") +
  scale_y_continuous(name="Northing (km)") +
  facet_wrap(~ design, ncol=1, nrow=2)+
  coord_fixed()
```

## Increasing the sample size {#IncreaseSampleSize}

Nine points are rather few for fitting a polynomial regression model, especially for a second-order polynomial with interaction. Therefore, in experiments often multiple observations are done for each design point. Increasing the sample size of a response surface sample in observational studies is not straightforward. The challenge is to avoid spatial clustering of sampling points. A simple solution is to select multiple points from each subset of candidate sampling units. The success of this solution depends on the strength of the spatial clustering of the candidate sampling units per design point. For the Cotton Research Farm for most design points the candidate sampling units are not in one spatial cluster, so in this case this solution may work properly. I increased the number of candidate sampling units per design point (argument `p` in function `quantile` is now 0.01), so that there is a larger choice in the optimisation of the sample pattern. 

```{r}
candi_all <- NULL
for (i in 1:nrow(ccd_df)) {
    d2dpnt <- sqrt((EM_CRF$PC1-ccd_df$x1[i])^2+
                     (EM_CRF$PC2-ccd_df$x2[i])^2)
    q <- quantile(d2dpnt, p=0.01)
    units_close <- which(d2dpnt < q)
    candi <- EM_CRF[units_close,c("unit","x1","x2","PC1","PC2")]
    candi$dpnt <- i
    candi_all <- rbind(candi_all,candi)
}
```

A stratified random subsample of two points per stratum is selected, which serves as an initial sample.

```{r}
set.seed(314)
units_stsi <- sampling::strata(
  candi_all, stratanames="dpnt", size=rep(2,9))
mysample0 <- getdata(candi_all, units_stsi)
mysample0 <- mysample0[,c("unit","x1","x2","PC1","PC2","dpnt")]
```

The `data.frame` with the design points must be doubled. Note that the order of the design points must be the equal to the order in the stratified subsample.

```{r}
tmp <- data.frame(ccd_df,dpnt=1:9)
dpnt <- rbind(tmp,tmp)
ccd_df2 <- dpnt[order(dpnt$dpnt),]
```

```{r, echo=FALSE, eval=FALSE}
set.seed(314)
mySRSsample <- anneal(
  mysample=mysample0,
  candidates=candi_all,
  dpnt=ccd_df2,
  phi=50,
  T_ini=1,
  coolingRate=0.9,
  maxPermuted=25*nrow(mysample0),
  maxNoChange=20,
  verbose=TRUE
)
save(mySRSsample, file="results/SpatialResponseSurfaceSample_2n_CRF_mb.RData")
```

Figures \@ref(fig:CCRSDUzbekistan2n) and \@ref(fig:CCRSDinPCSpace2n) show the optimised CCRSD sample of eighteen points in geographical and principal component space, respectively, obtained with the model-based criterion, an effective range of 50 m, and zero weight for the penalty term. Sampling points are not spatially clustered, so I do not expect violation of the assumption of independent residuals. In principal component space all points are pretty close to the design points, except for the four design points in the lower right corner, where no candidate units near these design points are available.

```{r CCRSDUzbekistan2n, echo=FALSE, fig.cap="CCRSD sample with two points per design point, from the Cotton Research Farm in Uzbekistan."}
load(file="results/SpatialResponseSurfaceSample_2n_CRF_mb.RData")
mysample<-mySRSsample$mysample
ggplot(EM_CRF) +
  geom_raster(aes(x=x1/1000,y=x2/1000,fill=PC1)) +
  geom_point(data=mysample,mapping=aes(x=x1/1000,y=x2/1000),size=2) +
  scale_fill_continuous(name="PC1",type = "viridis") + 
  scale_x_continuous(name="Easting (km)") +
  scale_y_continuous(name="Northing (km)") +
  coord_fixed()
```

```{r CCRSDinPCSpace2n, echo=FALSE, out.width='50%', fig.cap="CCRSD sample (triangles) with two points per design point (dots), optimised with model-based criterion."}
load(file="results/SpatialResponseSurfaceSample_2n_CRF_mb.RData")
mysample<-mySRSsample$mysample
ggplot(mysample) +
 geom_point(data=ccd_df,aes(x=x1,y=x2),size=3) +
 geom_point(aes(x=PC1,y=PC2),size=3,shape=2) +
 scale_x_continuous(name="PC1") +
 scale_y_continuous(name="PC2") +
 coord_fixed()
```

## Stratified spatial response surface sampling

The sample size can also be increased by stratified spatial response surface sampling\index{Stratified spatial response surface sampling}. The strata are subareas of the study area. When the subsets of candidate sampling units for some design points are strongly spatially clustered, the final optimised sample obtained with the method of the previous section may also show strong spatial clustering. An alternative is then to split the study area into two or more subareas (strata), and to select from each stratum candidate sampling units. This guarantees that for each design point we have at least as many spatial clusters of candidate units as we have strata. The spatial strata are not used for fitting separate regression models; all data are used to fit one (second-order) polynomial regression model. 

Figure \@ref(fig:StrataCRF4CCRSD) shows two subareas used as strata in stratified response surface sampling of the Cotton Research Farm.

```{r StrataCRF4CCRSD, echo=FALSE, fig.cap="Two subareas of the Cotton Research Farm in Uzbekistan used as strata in stratified CCRSD sampling."}
load(file="results/SpatialResponseSurfaceSample_2n_CRF_mb.RData")
ggplot(EM_CRF) +
  geom_raster(aes(x=x1/1000,y=x2/1000, fill=as.factor(subarea))) +
  scale_fill_grey(name="Strata") +
  scale_x_continuous(name="Easting (km)") +
  scale_y_continuous(name="Northing (km)") +
  coord_fixed()
```

The candidate sampling units are selected in a double for-loop. The outer loop is over the strata, the inner loop over the design points. Note that the variable `dpnt` continues to increase by 1 after the inner-loop over the nine design points in subarea 1 is completed, so that the variable `dpnt` (used as a stratum in subsampling the sample of candidate sampling points) now has values $1,2, \dots , 18$.  Stratum 2 is larger than stratum 1. To select an equal number of candidate sampling points per design point in both strata (eight points), I adapted argument `p` of function `quantile'. 

```{r}
candi_all <- NULL
p <- c(0.0105,0.009)
for (h in c(1,2)) {
  units <- which(EM_CRF$subarea==h)
  data_stratum <- EM_CRF[units,]
  candi_stratum <- NULL
  for (i in 1:nrow(ccd_df)) {
      d2dpnt <- sqrt((data_stratum$PC1-ccd_df$x1[i])^2+
                       (data_stratum$PC2-ccd_df$x2[i])^2)
      q <- quantile(d2dpnt,p=p[h])
      units_close <- which(d2dpnt < q)
      candi <- data_stratum[units_close,
                            c("unit","x1","x2","PC1","PC2")]
      candi$dpnt <- i + (h-1)*nrow(ccd_df)
      candi_stratum <- rbind(candi_stratum, candi)
  }
  candi_all <- rbind(candi_all, candi_stratum)
}
```

```{r, echo=FALSE, fig.cap="Candidate sampling points for stratified CCRSD sampling, plotted on first principal component (PC1)"}
subarea <- as.factor(c(rep(1,nrow(candi_all)/2),rep(2,nrow(candi_all)/2)))
dpnt9 <- c(rep(1:9, each=8), rep(1:9, each=8))
ggplot(EM_CRF) +
  geom_raster(aes(x=x1/1000,y=x2/1000,fill=PC1)) +
  geom_point(data=candi_all,mapping=aes(x=x1/1000,y=x2/1000,shape=as.factor(dpnt9)),size=1.5)+
  scale_shape_manual(values=c(0,1,2,3,4,5,6,7,8),name="design point") +
  scale_fill_continuous(name="PC1",type = "viridis") + 
  scale_x_continuous(name="Easting (km)") +
  scale_y_continuous(name="Northing (km)") +
  coord_fixed()
```

As before, `dpnt` is used as a stratum identifier to subsample the candidate sampling units. Finally, the number of rows in the `data.frame` `ccd_df` with the design points is doubled.

```{r}
set.seed(314)
units_stsi <- sampling::strata(
  candi_all, stratanames="dpnt", size=rep(1,18))
mysample0 <- getdata(candi_all, units_stsi)
mysample0 <- mysample0[,c("unit","x1","x2","PC1","PC2","dpnt")]
ccd_df2 <- rbind(ccd_df,ccd_df)
```

```{r, echo=FALSE, eval=FALSE}
set.seed(314)
mySRSsample <- anneal(
  mysample=mysample0,
  candidates=candi_all,
  dpnt=ccd_df2,
#  weight=5,
  phi=50,
  T_ini=1,
  coolingRate=0.9,
  maxPermuted=25*nrow(mysample0),
  maxNoChange=20,
  verbose=TRUE
)
save(mySRSsample, file="results/StratifiedSpatialResponseSurfaceSample_CRF_mb.RData")
#save(mySRSsample, file="results/StratifiedSpatialResponseSurfaceSample_CRF_mb_w5.RData")
```

Figures \@ref(fig:StratifiedCCRSD) and \@ref(fig:StratifiedCCRSDinPCSpace) show the optimised sample of eighteen points in geographical and principal component space, obtained with the model-based criterion with an effective range of 50 m. The pattern in the principal component space is worse compared to the pattern in Figure \@ref(fig:CCRSDinPCSpace2n). In stratum 1 the distance to the star point at the top and the upper left cube point is very large. In this stratum no population units are present that are close to these two design points. The distance could be decreased a bit by adding a penalty term to the minimisation criterion that is proportional to this distance. However, the optimised sample obtained with `weight=5` shows much stronger spatial clustering. In addition, there are some pairs of points with a separation distance smaller than the effective range used in optimisation, so that violation of the independence assumption becomes more likely. 

```{r StratifiedCCRSD, echo=FALSE, fig.cap="Stratified CCRSD samples from the Cotton Research Farm in Uzbekistan, optimised with model-based criterion, obtained without (weight = 0) and with penalty (weight = 5) for large average distance to design points."}
load(file="results/StratifiedSpatialResponseSurfaceSample_CRF_mb.RData")
mysample_w0 <- mySRSsample$mysample

load(file="results/StratifiedSpatialResponseSurfaceSample_CRF_mb_w5.RData")
mysample_w5 <- mySRSsample$mysample
mysamples <- rbind(mysample_w0,mysample_w5)
mysamples$weight <- rep(c("weight: 0","weight: 5"), each=18)

ggplot(mysamples) +
  geom_raster(data=EM_CRF, mapping=aes(x=x1/1000,y=x2/1000, fill=PC1)) +
  geom_point(data=mysample,mapping=aes(x=x1/1000,y=x2/1000), size=2) +
  scale_fill_continuous(name="PC1",type = "viridis") + 
  scale_x_continuous(name="Easting (km)") +
  scale_y_continuous(name="Northing (km)") +
  facet_wrap(~ weight, ncol=1, nrow=2) +
  coord_fixed()
```

```{r StratifiedCCRSDinPCSpace, echo=FALSE, out.width='100%', fig.cap="Principal component scores of the stratified CCRSD sample, optimised with model-based criterion, obtained without (weight = 0) and with penalty (weight = 5) for a large average distance to design points (dots)."}
load(file="results/StratifiedSpatialResponseSurfaceSample_CRF_mb.RData")
mysample_w0<-mySRSsample$mysample
mysample_w0$subarea <- as.factor(rep(c(1,2),each=9))

load(file="results/StratifiedSpatialResponseSurfaceSample_CRF_mb_w5.RData")
mysample_w5<-mySRSsample$mysample
mysample_w5$subarea <- as.factor(rep(c(1,2),each=9))

mysamples <- rbind(mysample_w0,mysample_w5)
mysamples$weight <- rep(c("weight: 0","weight: 5"), each=18)

ggplot(mysamples) +
 geom_point(data=ccd_df, mapping=aes(x=x1, y=x2), size=2) +
 geom_point(aes(x=PC1,y=PC2, shape=subarea), size=2) +
 scale_shape_manual(values=c(2,3), name="Stratum") +
 scale_x_continuous(name="PC1") +
 scale_y_continuous(name="PC2") +
 facet_wrap(~ weight) +
 coord_fixed()
```


```{r, echo=FALSE}
rm(list=ls())
```
