---
title: 'Exercise model-based optimisation of grid spacing: controlling a quantile of the ordinary kriging variance'
author: "Dick Brus"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r}
library(gstat)
library(ggplot2)
library(sp)
library(data.table)
library(tidyverse)
```

## Load data

Load the data on soil organic carbon of the three woredas in Ethiopia. Also load the discretisation grid of the study area.

```{r}
#load data on soil organic matter
load(file="../data/ThreeWoredasEthiopia.RData")
load(file="../data/CovariatesThreeWoredasEthiopia.RData")
```

## Estimation of the semivariogram

Use the data to compute a sample semivariogram and to fit a spherical model with nugget.

```{r}
#compute sample semivariogram
vg <- variogram(SOM~1, data = priordataEthiopia)
plot(vg)
#fit semivariogram model
vgm_MoM <- fit.variogram(vg, model = vgm(psill = 0.6, model = "Sph", range = 40, nugget = 0.6))
plot(vg,vgm_MoM)
print(vgm_MoM)
```

## Evaluation of grid spacings

### Select random sample of evaluation points

Select a simple random sample with replacement of size 1,000 for evaluating the square grids. Add a small number to the x-coordinates and y-coordinates by drawing from a uniform distribution with lower and upper limit equal to -cellsize/2 and +cellsize/2, respectively. This can be done with function `jitter`. The cell size is 1 km x 1 km

```{r}
set.seed(314)
units <- sample.int(nrow(grdEthiopia), size=1000, replace=TRUE)
mysample <- grdEthiopia[units,]

#Shift the randomly selected points to random points within the cells
mysample$s1 <- jitter(mysample$s1, amount=0.5)
mysample$s2 <- jitter(mysample$s2, amount=0.5)
```

### Kriging

Specify the grid spacings to be evaluated, and the cumulative probabilities from which the corresponding quantiles of the ordinary kriging variance at the 1,000 evaluation points are computed.

Select a square grid with the first spacing, add a dummy variable and predict the dummy variable at the evaluation points by ordinary kriging.  Repeat this for the other spacings (construct a for-loop). Finally plot the quantiles of the kriging variance against the grid spacing.

```{r}
coordinates(mysample) <- ~s1+s2
gridded(grdEthiopia) <- ~s1+s2

#define grid spacings
spacing <- 5:12

#specify cumulative probabilities
cprob <- c(0.5, 0.8, 0.95)

QKV <- matrix(nrow=length(spacing), ncol=length(cprob))
for (i in 1:length(spacing)) {
    mygridxy <- spsample(x=grdEthiopia, cellsize=spacing[i], type="regular", offset=c(0.5,0.5))
    #add a dummy variable for interpolation
    mygrid <- data.frame(s1=mygridxy$x1,s2=mygridxy$x2,dummy=1)
    coordinates(mygrid) <- ~s1+s2
    #Use gstat for ordinary kriging predictions
    predictions  <- krige(
        dummy ~ 1,
        locations = mygrid,
        newdata = mysample,
        model = vgm_MoM,
        nmax = 20
    )
    QKV[i,] <- quantile(predictions$var1.var, probs=cprob)
}

result <- data.table(spacing, QKV)
#save(result, file="../results/MBGridSpacing_Ethiopia_QOKV.RData")

df <- result %>% pivot_longer(.,cols=c("V1","V2","V3"))
ggplot(data=df)+
  geom_point(mapping=aes(x=spacing, y=value, shape=name), size=3)+
  scale_shape_manual(values=c(0,1,2), labels=c("P50","P80","P95"), name="Criterion") +
  scale_x_continuous(name="Spacing (km)")+
  scale_y_continuous(name="Quantile kriging Variance")
```
Finally, compute the tolerable grid-spacing so that P50, P80 and P95 of the ordinary kriging variance do not exceed 0.85

```{r}
names(result)[c(2,3,4)] <- c("P50","P80","P95")
(spacing_tol_P50 <- approx(x=result$P50, y=result$spacing, xout=0.85)$y)
(spacing_tol_P80 <- approx(x=result$P80, y=result$spacing, xout=0.85)$y)
(spacing_tol_P95 <- approx(x=result$P95, y=result$spacing, xout=0.85)$y)

```

