---
title: "Exercise sensititivity of model-based optimisation of grid-spacing to semivariogram parameters."
author: "Dick Brus"
date: "`r Sys.Date()`"
output: html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r}
library(sp)
library(gstat)
library(ggplot2)
library(tidyverse)
```

## Load data

Load the data on soil organic carbon of the three woredas in Ethiopia and a discretisation grid of the study area. 

```{r}
load(file="../data/ThreeWoredasEthiopia.RData")
load(file="../data/CovariatesThreeWoredasEthiopia.RData")
```

## Estimation of the semivariogram

Use the data to compute a sample semivariogram and to fit a spherical model with nugget by the method-of-moments.

```{r}
#compute sample semivariogram
vg <- variogram(SOM~1, data = priordataEthiopia)
#fit semivariogram model
vgm_MoM <- fit.variogram(vg, model = vgm(psill = 0.6, model = "Sph", range = 40, nugget = 0.6))
plot(vg, vgm_MoM)
print(vgm_MoM)
```

## Select random sample of evaluation points

Select a simple random sample of size 1,000 for evaluating the square grids. Add a small number to the x-coordinates and y-coordinates by drawing from a uniform distribution with lower and upper limit equal to -cellsize/2 and +cellsize/2, respectively. This can be done with function jitter. The cell size is 1 km x 1 km.

```{r}
set.seed(314)
units <- sample.int(nrow(grdEthiopia),size=1000,replace=TRUE)
mysample <- grdEthiopia[units,]

#Shift the randomly selected points to random points within the cells
mysample$s1 <- jitter(mysample$s1, amount=0.5)
mysample$s2 <- jitter(mysample$s2, amount=0.5)
```

## Kriging

Specify the grid spacings to be evaluated and the cumulative probabilities from which the corresponding quantiles of the ordinary kriging variance at the 1,000 evaluation points are computed. 

Select a square grid with the first spacing, add a dummy variable and predict the dummy variable at the evaluation points by ordinary kriging.  Repeat this for the other spacings (for-loop).

```{r}
coordinates(mysample) <- ~s1+s2
coordinates(grdEthiopia) <- ~s1+s2
gridded(grdEthiopia) <- TRUE

#define grid spacings
spacing <- 5:12

MKV <- NULL
for (i in 1:length(spacing)) {
    mygridxy <- spsample(x=grdEthiopia, cellsize=spacing[i], type="regular", offset=c(0.5,0.5))
    #add a dummy variable for prediction
    mygrid <- data.frame(s1=mygridxy$x1,s2=mygridxy$x2,dummy=1)
    coordinates(mygrid) <- ~s1+s2
    #Use gstat for ordinary kriging predictions
    predictions  <- krige(
        dummy ~ 1,
        mygrid,
        newdata = mysample,
        model = vgm_MoM,
        nmax = 20,
        debug.level = 0 
    )
    MKV[i] <- mean(predictions$var1.var)
}
```


## Exploring the sensitivity of grid spacing for semivariogram parameters

In practice we are uncertain about the semivariogram. For that reason it can be wise to explore the sensitivity of the determined grid spacing for the semivariogram parameters. 

### Sensitivity for nugget parameter

Increase the nugget by 5\%, and change the partial sill parameter so that the sill (nugget + partial sill) is equal to the original sill. 

```{r}
nugget <- vgm_MoM$psill[1] + 0.05*vgm_MoM$psill[1]
psill <- vgm_MoM$psill[1] + vgm_MoM$psill[2] - nugget
vgm_MoM_morenugget <- vgm(psill = psill, model = "Sph", range = vgm_MoM$range[2], nugget = nugget)

MKV_morenugget <- NULL
for (i in 1:length(spacing)) {
    mygridxy <- spsample(x=grdEthiopia,cellsize=spacing[i],type="regular",offset=c(0.5,0.5))
    mygrid <- data.frame(s1=mygridxy$x1,s2=mygridxy$x2,dummy=1)
    coordinates(mygrid) <- ~s1+s2
    predictions  <- krige(
        dummy ~ 1,
        mygrid,
        newdata = mysample,
        model = vgm_MoM_morenugget,
        nmax = 20,
        debug.level=0
    )
    MKV_morenugget[i] <- mean(predictions$var1.var)
}
```

### Sensitivity for range parameter

Do the same for a smaller range. Decrease the range by 5\%. Use the original values for nugget and partial sill.

```{r}
range <- vgm_MoM$range[2] - 0.05*vgm_MoM$range[2]
vgm_MoM_smallerrange <- vgm(psill = vgm_MoM$psill[2], model = "Sph", range = 30, nugget = vgm_MoM$psill[1])
MKV_smallerrange <- NULL
for (i in 1:length(spacing)) {
    mygridxy <- spsample(x=grdEthiopia,cellsize=spacing[i],type="regular",offset=c(0.5,0.5))
    mygrid <- data.frame(s1=mygridxy$x1,s2=mygridxy$x2,dummy=1)
    coordinates(mygrid) <- ~s1+s2
    predictions  <- krige(
        dummy ~ 1,
        mygrid,
        newdata = mysample,
        model = vgm_MoM_smallerrange,
        nmax = 20,
        debug.level=0
    )
    MKV_smallerrange[i] <- mean(predictions$var1.var)
}
```


### Tolerable grid spacing and required expected sample size

Plot the MKV's obtained  with the three semivariograms against the grid spacing. 

```{r}
df <- data.frame(spacing, MKV, MKV_morenugget, MKV_smallerrange)
#save(df,file="../results/MBGridSpacing_Ethiopia_Sensitivity.RData")

df_lf <- df %>% pivot_longer(.,cols=c("MKV","MKV_morenugget","MKV_smallerrange"))
ggplot(data=df_lf)+
  geom_point(mapping=aes(x=spacing, y=value, shape=name), size=3)+
  scale_shape_manual(values=c(0,1,2), labels=c("Original","More nugget","Smaller range"), name="Semivariogram") +
  scale_x_continuous(name="Spacing (km)")+
  scale_y_continuous(name="Mean kriging variance")
```

Compute the tolerable grid spacing and the corresponding expected sample size for a MKV value of 0.85.

```{r}
(spacings_tol <- lapply(1:3, function (i) approx(x=df[,i+1], y=df$spacing, xout=0.85)$y) %>% as.numeric(.))
#compute size of pixels
res2 <- getGridTopology(grdEthiopia) %>% data.frame(.)
#compute total area in km2
A <- length(grdEthiopia) * res2$cellsize[1] * res2$cellsize[2]
#compute required expected sample size
(nreq <- A/spacings_tol^2)

```

