---
title: "Exercise optimisation of grid spacing for KED, using the mean kriging variance as a criterion"
author: "Dick Brus"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r}
library(gstat)
library(sp)
library(ggplot2)
```


## Residual semivariogram

Define residual semivariogram.

```{r}
vgmresi_REML <- vgm(model = "Exp", psill = 0.083, range = 230, nugget = 0.126)
```


## Evaluation of grid spacings

### Select random sample of prediction points

Select a simple random sample with replacement of size 1000 for evaluating the square grids. Add a small number to the x1-coordinates and x2-coordinates by drawing from a uniform distribution with lower and upper limit equal to -cellsize/2 and +cellsize/2, respectively. The cell size is 25 m x 25 m.

```{r}
#read discretisation grid with interpolated natural log-transformed EM values
load(file="../data/CottonResearchFarm.RData")

#select simple random sample with replacement of size 1000 
set.seed(314)
units <- sample(nrow(EM_CRF), size=1000, replace=TRUE)
mysample <- EM_CRF[units,]

#Shift the randomly selected grid points to random points within the cells
mysample$x1 <- jitter(mysample$x1, amount=25/2)
mysample$x2 <- jitter(mysample$x2, amount=25/2)

coordinates(mysample) <- ~x1+x2
```

### Kriging with an external drift

Choose grid spacings to be evaluated. Take 100 m as the largest spacing. The farm is about 100 ha, so with this spacing the number of grid points is on average about 100. Take 70 m as the smallest spacing.

Select a randomly placed square grid of the first spacing, add a dummy variable and predict the dummy variable at the prediction points by kriging with an external drift, using lnEM100cm as a predictor. Compute the average of the kriging variances at the 1,000 prediction points, and save it in a matrix. Repeat this ten times (inner for-loop), and repeat this for the other spacings (outer for-loop).

```{r kriging}
#specify spacings
spacing <- seq(from=70, to=100, by=5)

#set number of times grid sampling of a given spacing is repeated
r <- 10

gridded(EM_CRF) <- ~x1+x2
MKV <- matrix(nrow=length(spacing),ncol=r)

for (i in 1:length(spacing)) {
    for (j in 1:r) {
        mygridxy <- spsample(x=EM_CRF, cellsize=spacing[i], type="regular")
        #add a dummy variable for interpolation
        mygrid <- data.frame(s1=mygridxy$x1, s2=mygridxy$x2, dummy=1)
        coordinates(mygrid) <- ~s1+s2
        mygrd <- data.frame(over(mygrid, EM_CRF), mygrid)
        coordinates(mygrd) <- ~s1+s2
        #Use gstat for KED predictions
        predictions  <- krige(
            dummy ~ lnEM100cm,
            mygrd,
            newdata = mysample,
            model = vgmresi_REML,
            nmax = 20,
            debug.level = 0
        )
        preds <- as.data.frame(predictions)
    MKV[i,j] <- mean(predictions$var1.var)
    }
}
```

## Results

Plot the mean kriging variance against the grid spacing.

```{r}
MMKV <- apply(MKV,MARGIN=1,FUN=mean)
df <- data.frame(spacing,MMKV)

ggplot(data=df)+
  geom_point(mapping=aes(x=spacing,y=MMKV),size=3)+
  scale_x_continuous(name="Spacing (m)")+
  scale_y_continuous(name="Mean kriging variance",limits=c(0.16,0.17))
```
Compute tolerable grid spacing for a MMKV of 0.165.

```{r}
print(spacing_tol <- approx(x=df$MMKV, y=df$spacing, xout=0.165)$y)
```

