---
title: "Exercise model-based optimisation of sampling locations for ordinary kriging"
author: "Dick Brus"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(sp)
library(spsann)
library(gstat)
library(ggplot2)
```

## Discretisation grid

Create a grid that discretises the population.
```{r}
s1 <- s2 <- 1:20 - 0.5
grid <- expand.grid(s1,s2)
names(grid) <- c("s1","s2")
```

## Semivariogram

The coordinates of sixteen sampling locations are optimised for ordinary kriging. First, the semivariogram is defined.

```{r}
nugget <- 0
partial_sill <- 2
range <- 4
vgm_gstat <- vgm(model="Exp", psill=partial_sill, range=range, nugget=nugget)
```

Specify the candidate sampling points. Note that not the grid points themselves are used as candidate points, but all points within the grid cells with the grid points at their centres. So, the set of candidate points is not finite but infinite. In the optimisation a new candidate point is selected in two steps. First, a grid cell is randomly selected. Then a random point within the selected grid cell is selected.

```{r}
candi <- grid[,c(1,2)]
names(candi) <- c("x","y") #these names are required by spsann 
```

## Optimising the sampling locations

### The simulated annealing schedule

Define the schedule for simulated annealing. Note that both the initial acceptance rate  and the initial temperature are set, which may seem weird as the acceptance rate is a function of the initial temperature: $P =e^{\frac{-\Delta f}{T}}$. The initial acceptance rate is used as a threshold value. If an initial temperature is chosen that leads to an acceptance rate smaller than the chosen value for the initial acceptance rate, then the optimisation stops. In this case a larger value for the initial temperature must be chosen.

The chain.length is the number of iterations during which the temperature is kept constant. The value for chain.length is a multiplier, see ?scheduleSPSANN. So, when for instance 50 locations are optimised, and a value 2 is chosen for chain.length, the chain length equals 50 x 2 = 100 iterations. 

```{r}
schedule <- scheduleSPSANN(initial.acceptance = 0.8,
                           initial.temperature = 0.07,
                           temperature.decrease=0.90,
                           chains=500,
                           chain.length=10,
                           stopping=10,
                           x.max=5,y.max=5,
                           cellsize=1)
```

### Optimisation of sampling locations

Now start the simulated annealing algorithm.

```{r}
set.seed(314)
res <- optimMKV(
  points = 16, 
  candi = candi, 
  vgm = vgm_gstat,
  eqn = z ~ 1, 
  schedule = schedule,
  track=TRUE)
#save(res,file = "../results/MBSampleSquare_OK_NoNugget_16pnts.RData")
```

## Result

Plot the trace to check convergence

```{r}
#load(file = "../results/MBSampleSquare_OK_NoNugget_16pnts.RData")
trace <- res$objective$energy
ggplot(trace) +
  geom_line(mapping = aes(x=1:nrow(trace), y = obj), colour="red") +
  scale_y_continuous(name = "Mean kriging variance") +
  scale_x_continuous(name = "Iteration")
```

Plot the optimised sample

```{r}
sample <- res$points
ggplot(data = grid) +
  geom_tile(mapping = aes(x = s1, y = s2), fill="grey") +
  geom_point(data=sample,mapping = aes(x = x, y = y), size=2) +
  geom_vline(xintercept=c(5,10,15)) +
  geom_hline(yintercept=c(5,10,15)) +
  scale_y_continuous(name = "Northing") +
  scale_x_continuous(name = "Easting") +
  coord_fixed()
```

