---
title: "Exercise optimisation of sampling pattern for ordinary kriging, using the P90 of the ordinary kriging variance as a criterion"
author: "Dick Brus"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r}
library(gstat)
library(sp)
library(spsann)
library(ggplot2)
```


## Semivariogram

The parameters of an exponential semivariogram are specified. These parameters are shown in Table 23.1 
```{r}
vgm_lnECe <- vgm(model = "Exp", psill = 0.336, range = 192, nugget = 0.116)
```

## Read discretisation grid

Read square grid that discretises the study area. This 25 x 25 grid serves as the evaluation sample, and is also used to select candidate sampling points.

```{r}
load(file="data/CottonResearchFarm.RData")
mysample_eval <- EM_CRF[,c("x1","x2")]
names(mysample_eval) <- c("x","y")
candi <- mysample_eval
```

## Objective function

Define function to compute objective function (P90 of ordinary kriging variance) for a given sample.

```{r}
QOKV <- function(points, esample, model, nmax, prob) {
  points <- as.data.frame(points)
  coordinates(points) <- ~x+y
  # add dummy variable
  points$dum=1
  # compute variance of prediction error
  res <- krige(
    formula = dum~1,
    locations = points,
    newdata = esample,
    model = model,
    nmax = nmax,
    debug.level = 0
  )
  quantile(res$var1.var, probs=prob)
}
```

## Optimisation of sampling pattern through simulated annealing

Define the simulated annealing schedule.

```{r}
#Specify annealing schedule
schedule <- scheduleSPSANN(
  initial.acceptance = 0.8,
  initial.temperature = 0.005,
  temperature.decrease = 0.9,
  chains = 100,
  chain.length = 2,
  stopping = 10,
  x.min = 25, y.min = 25,
  cellsize = 25)
```


Start the optimisation through annealing.

```{r}
coordinates(mysample_eval) <- ~x+y

set.seed(314)
rslt <- optimUSER(
  points = 50,
  candi = candi,
  fun = QOKV,
  esample = mysample_eval,
  model = vgm_lnECe,
  nmax = 20,
  prob = 0.9,
  schedule = schedule,
  track = TRUE)

mysample <- candi[rslt$points$id,]
save(mysample, file="../results/MBSample_OK_P90_Uzbekistan.RData")
```

## Result

```{r}
ggplot(data=EM_CRF) +
  geom_raster(mapping = aes(x = x1/1000, y = x2/1000),fill="grey")+
  geom_point(data = mysample, mapping = aes(x = x/1000, y = y/1000), shape = 2,size= 2)+
  scale_x_continuous(name = "Easting (km)")+
  scale_y_continuous(name = "Northing (km)")+
  coord_fixed() 
```

