---
title: "Exercise optimisation of sampling pattern for ordinary kriging, using the P90 of the ordinary kriging variance as a minimisation criterion"
author: "Dick Brus"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r}
library(gstat)
library(sp)
library(spsann)
library(ggplot2)
library(sswr)
```


## Semivariogram

Specify the parameters of an exponential semivariogram for lnECe.

```{r}
vgm_lnECe <- vgm(model = "Exp", psill = 0.336, range = 192, nugget = 0.116)
```

## Define evaluation sample and candidate sampling points 

The 25 m $\times$ 25 m grid `grdCRF` serves as the evaluation sample, and is also used to select candidate sampling points.

```{r}
mysample_eval <- grdCRF[, c("x", "y")]
candi <- mysample_eval
```

## Objective function

Define the objective function (P90 of ordinary kriging variance) for a given sample.

```{r}
QOKV <- function(points, esample, model, nmax, prob) {
  points <- as.data.frame(points)
  coordinates(points) <- ~ x + y
  # add dummy variable
  points$dum <- 1
  # compute variance of prediction error
  res <- krige(
    formula = dum ~ 1,
    locations = points,
    newdata = esample,
    model = model,
    nmax = nmax,
    debug.level = 0
  )
  quantile(res$var1.var, probs = prob)
}
```

## Optimisation of sampling pattern through simulated annealing

Define the simulated annealing schedule. Use an initial acceptance rate a range from 0.8 to 0.95, a temperature decrease of 0.9, a maximum of 500 chains, `chain.length = 2` (number of iterations per chain is then two times the sample size), a stopping criterion of 10 chains, and `cellsize = 0` so that the sampling points are restricted to the centres of the 25 m $\times$ 25 m grid. Choose an initial temperature so that the acceptance rate equals 0.8 or larger. 

```{r}
#Specify annealing schedule
schedule <- scheduleSPSANN(
  initial.acceptance = c(0.8, 0.95),
  initial.temperature = 0.006,
  temperature.decrease = 0.9,
  chains = 500,
  chain.length = 2,
  stopping = 10,
  cellsize = 25)
```


Optimise the locations of 50 points. Use a maximum of 20 points in ordinary kriging.. 

```{r}
coordinates(mysample_eval) <- ~ x + y

set.seed(314)
rslt <- optimUSER(
  points = 50,
  candi = candi,
  fun = QOKV,
  esample = mysample_eval,
  model = vgm_lnECe,
  nmax = 20,
  prob = 0.9,
  schedule = schedule,
  track = TRUE)

mysample <- rslt$points
write_rds(mysample, file = "../results/MBSample_OK_P90_Uzbekistan.rds")
```

## Result

Plot the optimised sample

```{r}
ggplot(data = grdCRF) +
  geom_raster(mapping = aes(x = x / 1000, y = y / 1000), fill = "grey") +
  geom_point(data = mysample, mapping = aes(x = x / 1000, y = y / 1000), shape = 2, size = 2) +
  scale_x_continuous(name = "Easting (km)") +
  scale_y_continuous(name = "Northing (km)") +
  coord_fixed()
```

